# [5.2 Plotting geographci data](https://walker-data.com/census-r/census-geographic-data-and-applications-in-r.html#plotting-geographic-data)

In the previous section, we have used the `plot()` function to visualize the 
geometry column of a simple features object obtained with **tigris**. 
R includes a variety of other options for quick visualization of geographic data 
that will be useful to geospatial analysts.

## 5.2.1 ggplot2 and `geom_sf()`
As of ggplot2 version 3.0, the package released support for plotting simple 
features objects directly with the function `geom_sf()`. `geom_sf()` streamlines
the geographic visualization process and makes **ggplot2** a go-to package
for visualization of simple features objects.

At a basic level, a couple lines of **ggplot2** code are all that are needed to
plto Census shapes obtained with **tigris**.

```{r}
library(ggplot2)

la_tracts <- tracts("NM", "Los Alamos")

ggplot(la_tracts) + geom_sf()
```

By default, ggplot2 includes its standard grey grid with latitude and longitude values displayed along the axes. For many cartographic applications, an analyst will want to remove this background information. The theme_void() function strips the background grid and axis labels from the plot accordingly:

```{r}
ggplot(la_tracts) + 
  geom_sf() +
  theme_void()
```

For comparative spatial plots, the [patchwork R package](https://github.com/thomasp85/patchwork) works very well for arranging a multi-plot layout. Below, we’ll use patchwork to put two ggplot2 spatial plots - one of Census tracts and one of block groups in Los Alamos County - side-by-side using the + operator.

```{r}
library(patchwork)

la_block_groups <- block_groups("NM", "Los Alamos")

gg_tracts <- ggplot(la_tracts) +
  geom_sf() +
  theme_void() +
  labs(title = "Census tracts")

gg_blocks <- ggplot(la_block_groups) +
  geom_sf() + 
  theme_void() +
  labs(title = "Block groups")

gg_tracts + gg_blocks
```
Alternatively, **patchwork** allows R users to arrange plots vertically with the `/` operator.

## 5.2.2 Interactive viewing with mapview

The [mapview](https://r-spatial.github.io/mapview/) R package is used for interactive mapping. 
With a single call to its function `mapview()`, mapview visualizes geographic data 
on an interactive, zoomable map.

```{r}
library(mapview)

mapview(la_tracts)
```

Clicking on a Census tract shape reveals a pop-up with attribute information 
found in the dataset. Additionally, users can change the underlying basemap to 
understand the geographic context surrounding their data. **mapview** also includes 
significant functionality for interactive mapping and data display beyond this 
basic example; some of these features are covered in the next chapter.


