# [Grouped data](https://dplyr.tidyverse.org/articles/grouping.html)

dplyr verbs are particularly powerful when you apply them to grouped data frames (`grouped_df` objects). This vignette shows you:

* How to group, inspect, and ungroup with `group_by()` and friends.
* How individual dplyr verbs changes their behavior when applied to grouped data frame.
* How to access data about the "current" group from within a verb.

We'll start by loading dplyr:

```{r} library(dplyr)```

## `group_by()`

The most important is `group_by()`: it takes a data frame and one or more variables to group by:

```{r}
by_species <- starwars %>% group_by(species)
by_sex_gender <- starwars %>% group_by(sex, gender)
```

You can see the grouping when you print the data:
```{r}
by_species
by_sex_gender
```

Or use `tally()` to count the number of rows in each group. The `sort` arguement is useful if you want to see the largest groups up front.

```{r}
by_species %>% tally()
by_sex_gender %>% tally(sort = TRUE)
```

As well as grouping by existing variables, you can group by any function of existing variables. This is equivalent to performing a `mutate()` **before** the `group_by()`:

```{r}
bmi_breaks <- c(0, 18.5, 25, 30, Inf)

starwars %>%
    group_by(bmi_cat = cut(mass / (height / 100)^2, breaks=bmi_breaks)) %>%
    tally()
```
## Group metadata

You can see underlying group data with `group_keys()`. It has one row for each group and one column for each grouping variable.

```{r}
by_species %>% group_keys()
by_sex_gender %>% group_keys()
```

You can see which group each row belongs to with `group_indices()`:

```{r} by_species %>% group_indices()```

And which rows each group contains with `group_rows()`:

```{r} by_species %>% group_rows() %>% head()```

Use `group_vars()` if you just want the names of the grouping variables:

```{r}
by_species %>% group_vars()
by_sex_gender %>% group_vars()
```

### Changing and adding to grouping variables

If you apply `group_by()` to an already grouped dataset, it will overwrite the existing grouping variables. For example, the following code groups by `homeworld` instead of `species`:

```{r}
by_species %>%
    group_by(homeworld) %>%
    tally()
```
To **augment** the grouping, using `.add = TRUE`. For example, the following code groups by species and homeworld:

```{r}
by_species %>%
    group_by(homeworld, .add = TRUE) %>%
    tally()
```

### Removing grouping variables

To remove all grouping variables, use `ungroup()`:

```{r}
by_species %>%
    ungroup() %>%
    tally()
```
You can also choose to selectively ungroup by listing the variables you want to remove:

```{r}
by_sex_gender %>%
    ungroup(sex) %>%
    tally()
```

## Verbs

The following sections describe how grouping affects the main dplyr verbs.

### `summarize()`

`summarize()` computes a summary for each group. This means that it starts from `group_keys()`, adding summary variables to the right hand side:

```{r}
by_species %>%
    summarize(
        n = n(),
        height = mean(height, na.rm = TRUE)
    )
```

The `.groups=` arguement controls the grouping structure of the output. The historical behavior of removing the right hand side grouping variable corresponds to `.groups = "drop_last"` without a message or `.groups = NULL` with a message (the default).

```{r}
by_sex_gender %>%
    summarize(n = n()) %>%
    group_vars()

by_sex_gender %>%
    summarize(n = n(), .groups = "drop_last") %>%
    group_vars()
```

Since version 1.0.0 the groups may also be kept (`.groups = "keep"`) or dropped (`.groups = "drop"`)

```{r}
by_sex_gender %>%
    summarize(n = n(), .groups = "keep") %>%
    group_vars()

by_sex_gender %>%
    summarize(n = n(), .groups = "drop") %>%
    group_vars()
```

When the output no longer have grouping variables, it becomes ungrouped (i.e. a regular tibble).

### `select()`, `rename()`, and `relocate()`

`rename()` and `relocate()` behave identically with grouped and ungrouped data because they only affect the name or position of existing columns. Grouped `select()` is almost identical to ungrouped select, except that it always includes the grouping variables:

```{r} by_species %>% select(mass)```

If you don't want the grouping variables, you'll have to first `ungroup()`. (This design is possibly a mistake, but we're stuck with it for now.)

### `arrange()`

Grouped `arrange()` is the same as ungrouped `arrange()`, unless you set `.by_group = TRUE`, in which case it will order first by the grouping variables.

```{r}
by_species %>%
    arrange(desc(mass)) %>%
    relocate(species, mass)

by_species %>%
    arrange(desc(mass), .by_group = TRUE) %>%
    relocate(species, mass)
```

Note that second example is sorted by `species` (from the `group_by()` statement) and then by `mass` (within species).

### `mutate()`

In simple cases with vectorized functions, grouped and ungrouped `mutate()` give the same results. They differ when used with summary functions:

```{r}
# Subtract off global mean
starwars %>%
    select(name, homeworld, mass) %>%
    mutate(standard_mass = mass - mean(mass, na.rm = TRUE))

# Subtract off homeworld mean
starwars %>%
    select(name, homeworld, mass) %>%
    group_by(homeworld) %>%
    mutate(standard_mass = mass - mean(mass, na.rm = TRUE))
```

Or with window functions like `min_rank()`:

```{r}
# Overall rank
starwars %>%
    select(name, homeworld, height) %>%
    mutate(rank = min_rank(height))

# Rank per homeworld
starwars %>%
    select(name, homeworld, height) %>%
    group_by(homeworld) %>%
    mutate(rank = min_rank(height))
```

### `filter()`

A grouped `filter()` effectively does a `mutate()` to generate a logical variable, and then only keeps the rows where the variable is `TRUE`. This means that grouped filters can be used with summary functions. For example, we can find the tallest character of each species:

```{r}
by_species %>%
    select(name, species, height) %>%
    filter(height == max(height))
```

You can also use `filter()` to remove entire groups. For example, the following code eliminates all groups that only have a single member:

```{r}
by_species %>%
    filter(n() != 1) %>%
    tally()
```

### `slice()` and friends
`slice()` and friends (`slice_head()`, `slice_tail()`, `slice_sample()`, `slice_min()`, and `slice_max()`) select rows within a group. For example, we can select the first observation within each species:

```{r}
by_species %>%
    relocate(species) %>%
    slice(1)
```

Similarly, we can use `slice_min()` to select the smallest `n` values of a variable:

```{r}
by_species %>%
    filter(!is.na(height)) %>%
    slice_min(height, n = 2)
```