# Your First Plot!

Last week, you tracked how many cups of coffee you drank each day. Create a plot to show how your coffee intake varied over the week.

-   Initialize a ggplot and pass the `coffee_counter` data.
-   Map the variables to the correct positional aesthetics: the day of the week on the x-axis and the number of cups on the y-axis.
-   Plot the values using the `geom_point()` function.
-   Update the data to match your own coffee consumption

```{r}
library(ggplot2)

# Store some data on coffee consumption
coffee_counter <- data.frame(
  # day of the week
  wday = 1:7,
  # number of cups
  cups = c(2, 2, 1, 3, 2, 2, 2)
)

# Create the plot (your task!)
ggplot(data = coffee_counter, 
              aes(x = wday, y = cups)) +
geom_point()
```

# Connecting the Dots

Your coffee plot shows how much you drank each day — but right now, the dots are floating without any clear connection. Let’s fix that by **adding** a line to **highlight the progression** throughout the week.

-   Start with your previous `geom_point()` plot.
-   Add another layer to draw a line that connects the points.

```{r}
# Create the plot
ggplot(data = coffee_counter) +
  aes(x = wday, y = cups) +
  geom_point() +
  geom_line()
```

# Switching Between Chart Types

Our current plot connects the dots to show how many cups were consumed across days — but a bar chart might be a better fit to highlight totals.

Let’s see how the same data feels with a different geometry! **Replace the current layer with `geom_col()`** to map the number of cups (y aesthetic) to the height of the bars instead.

```{r}
# Create the plot
ggplot(data = coffee_counter) +
  aes(x = wday, y = cups) +
  geom_col()
```
# Changing the Theme
Want a cleaner look for your chart? Use themes to refine the overall appearance.

* Try `theme_minimal()` for a simpler, more modern look.
* Use the `base_size` argument to adjust the font size.

```{r}
# Create the plot
ggplot(data = coffee_counter) +
  aes(x = wday, y = cups) +
  geom_col() +
  theme_minimal(base_size = 20) # Default base_size is 11
```
# Modifying Data Types
The previous charts are great, but the x-axis is not really self-explanatory.

Instead of weekday numbers (`1` → `7`), we want to show **explicit weekday names** like `Mon` or `Tue.`

* How could we adjust the data to get those labels to show up in the plot? Think about changing the type of the x variable in your data.

```{r}
# To display the weekdays as labels like "Mon"
# or "Tue", we need to turn the numeric weekday
# variable into a factor with the correct order.
# This tells ggplot2 to treat it as categorical
# data and use a discrete scale on the x-axis.

coffee_counter <- data.frame(
  wday = factor(format(ISOdate(1, 1, 1:7), "%a"), levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")), 
  cups = c(3, 4, 5, 4, 6, 1, 2)
)

ggplot(data = coffee_counter) +
  aes(x = wday, y = cups) +
  geom_col()
```

