# Encoding Categories

Let's **add color to distinguish between drivetrain** types and uncover those patterns!

- Explore the dataset variables with `names(mpg)`
- Add the `color` aesthetic to map the drivetrain variable
- Observe how the color reveals group-specific trends
```{r}
library(ggplot2)

names(mpg)
```

```{r}
# Scatter plot of engine size vs city mileage
# Your task: Map color to drivetrain types
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point(aes(color = drv))

# Color helps us see group patterns:
# - 4-wheel drive cars (red) tend to cluster
#   with larger engines and lower mpg
# - Front-wheel drive cars (green) often
#   have smaller engines and better mpg
# - Rear-wheel drive cars (cyan) are more
#   spread out with larger engines
```

# Alternatives to Provide Aesthetics
There are two other ways to specify the aesthetics that would produce the same result as the plot above. Can you find them?

Hint: Aesthetics can be defined in their own aes() call—either inside ggplot() or at the layer level.

```{r}
# aes in the ggplot()
ggplot(mpg, aes(x = displ, y = cty, color = drv)) +
  geom_point()

# aes on it's own
ggplot(mpg) +
  aes(x = displ, y = cty, color = drv) +
  geom_point()

# aes in the geom layer
ggplot(mpg) +
  geom_point(aes(x = displ, y = cty, color = drv))
```
# Mapping Multiple Aesthetics
The previous plot used color to distinguish `drv`, which was great for spotting groups. Now, **add shapes to the points for `drv` as well** to make it easier for everyone to tell them apart.

Tip: Increase the point size by passing a number to the `size` argument inside `geom_point()` so the shapes are more visible.

Using multiple aesthetics (like color and shape) to encode the same variable is called _double encoding_. It can _improve accessibility_ and help viewers spot patterns more easily.

```{r}
# Double encode drv with color and shape!
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point(
    aes(color = drv, shape = drv), size = 3
  )

# If you look closely, you'll see points
# with unexpected shape-color combinations:
# This happens because of overplotting,
# which can be confusing at first!
```

# Mapping for Multiple Layers
To explore the trends in the data for the different drivetrain types, you're asked to **add regression lines** to your plot. Your colleague told you to replace the default method by `method = "lm"` to draw linear smoothings.

Yay, that works 🎊 But wait — _you only get a single line?!_

Fix the code below to show one line per drivetrain
What happens if you map the `drv` column to `shape` inside `geom_smooth()`? And why?

When using more than a single layer, the difference between global and local mapping is crucial! While global mappings are passed on to the layers, local mapping only applies to the individual layer.

```{r}
# Your Task: Draw one fitting per drv type
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point(
    aes(color = drv, shape = drv),
    size = 3
  ) +
  geom_smooth(method = "lm")

# Solution 1: Repeat color mapping locally
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point(
    aes(color = drv, shape = drv),
    size = 3
  ) +
  geom_smooth(aes(color = drv), method = "lm")

# Solution 2:
# Linear fitting for all three drv types
# -> global mapping of color (and shape)
ggplot(mpg, aes(x = displ, y = cty,
                color = drv, shape = drv)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm")
```

# Colorful Bars
The barplot below shows how many data points are available for each continent in the `hyde` dataset. 

Many people enjoy adding colors to their charts. How would you give each bar a different color here?

Hint: Like bars, certain point shapes let you control both the inside and the outline color. To change the inside, **you’ll need the right aesthetic**.

Adding color can be fun, but take a moment — **is it actually helping tell the story here**?

It’s common to see bar plots where each bar has a different color. But that’s rarely helpful. Using many colors without a clear reason can distract from your main message — or even confuse your audience. It’s best to use color thoughtfully — only when it clearly adds meaning.

```{r}
# source HYDE data from URL
hyde <- read.csv("https://www.ggplot2-uncharted.com/data/hyde.csv")

ggplot(hyde, aes(x = continent)) +
  geom_bar(aes(fill = continent)) # Adding color to the fill aesthetic here is unnecessary
```
# Highlight Categories
This bubble chart shows _urban population versus the logarithmic GDP per capita, with bubble size indicating each country’s total population_.

But it feels a bit... flat. It just shows the data, without telling a story.

**Let’s change that: highlight Asian countries in red!**
(As before, the `names()` function might come handy to find the right column.)

Hint: Some point shapes — like the one used here — let you control both the inside and the outline color. To change the inside, _you’ll need the right aesthetic_. You can map logical expressions directly inside `aes()` — a neat way to highlight parts of your data on the fly, without mutating your dataset.

```{r}
names(hyde)
```


```{r}
ggplot(
    hyde,
    aes(x = log(gdp_per_capita),
        y = urban_pop)) +
  geom_point(
    aes(size = pop_est, fill = (continent != "Asia")),
    alpha = 0.7, shape = 21
  )
```


