Bars, lines, points, and beyond — geometric layers define how your data appears on your panel. Think of them as the “verbs” in your visualization sentence.

This lesson explores how to pick the right geom for the job and how different layers can be combined to tell richer, more nuanced stories.

# The Final Ingredient

In the previous lessons, we covered the plot’s backbone — your data, and the aesthetics that map it to visual properties. But if you tried running a plot with just those, you’ll have noticed: nothing shows up inside the panel (a.k.a. the gray area).

```{r}
library(ggplot2)

ggplot(mpg) + aes(x = displ, y = cty)
```

That’s because every ggplot needs at least one layer to actually display anything. And most often, that layer is a geom\_\*().

So far, you’ve already seen and used several of these layers:

`geom_point()` → show individual observations `geom_line()` → connect data points `geom_col()` → display values as bars `geom_smooth()` → add trend lines `geom_text()` → label points with text

You might have spotted a few more, like the bar and segment geoms, or the reference-line trio with abline, hline & vline.

# Not All Geoms Want the Same Things

All of the geoms listed above will happily show up when added to our "empty panel" code with just `x` and `y` mapped — well, all but one.

```{r}
# Empty
ggplot(mpg) +
aes(x = displ, y = cty)

# Point
ggplot(mpg) +
aes(x = displ, y = cty) + 
  geom_point()

# Line
ggplot(mpg) +
aes(x = displ, y = cty) + 
  geom_line()

# Column
ggplot(mpg) +
aes(x = displ, y = cty) + 
  geom_col()

# Smooth
ggplot(mpg) +
aes(x = displ, y = cty) + 
  geom_smooth()

# Text?
ggplot(mpg) +
aes(x = displ, y = cty) + 
  geom_text()
```

It’s `geom_text()` that stays silent. Not because of a bug — but because it’s still waiting for one more piece: a `label` aesthetic. While we can draw a point or bar with two coordinates, `geom_text()` and `geom_label()` need to know **what** to write.

This isn’t unique to `geom_text()`. For example, `geom_point()` can’t draw a point without knowing where to place it, so it expects both `x` and `y`. Otherwise ggplot2 will throw an error and inform you about missing aesthetics:

```         
Error in `geom_point()`: Problem while setting up geom. `geom_point()` requires the following missing aesthetics: y.
```

*Every layer expects certain aesthetics to be present*:

-   Geometries that require two positions, usually `x` and `y`, are the most common case.
-   Others need a third aesthetic such as `label` for `geom_text()` or `fill` for `geom_tile()`.
-   Some layers are minimalists: for example, `geom_bar()` only required an `x` or `y` position to place the bar — and calculates the height or width, respectively, itself.
-   And some demand many more, often specific aesthetics such as additional `xend` and `yend` to draw a line with `geom_segment()` to define start and end coordinates.

Now that you know some geoms need different numbers and types of aesthetics, let’s explore them in more detail.

# A World of ~~Chart Types~~ Geoms

The world of data visualization includes countless chart types. While the possibilities are virtually endless, mastering the fundamental ones is key to clear and effective communication.

Below is a quick overview of chart types, grouped by their main purpose. ggplot2 can create all of them → it's just a matter of choosing the right geometric layer.

## Distribution

1.  Violin: `geom_violin()`
2.  Density: `geom_density()`
3.  Histogram: `geom_histogram()`
4.  Boxplot: `geom_boxplot()`
5.  Ridgeline: `{ggridges}` package

## Correlation

1.  Scatterplot: `geom_point()`
2.  Heatmap: `geom_tile()`
3.  Correlogram: `{GGally}` package
4.  Bubble: `geom_point()`
5.  2d Density: `geom_density_2d()`
6.  Connected scatter: `geom_point()` and `geom_path()`

## Ranking

1.  Barplot: `geom_bar()`
2.  Wordcloud: `{wordcloud}` and `{wordcloud2}` package
3.  [Parallel](https://www.data-to-viz.com/graph/parallel.html): `{GGally}` package
4.  Lollipop: `geom_point()`
5.  Circular barplot: `geom_bar()`
6.  Spider/Radar: `geom_line()`

## Part of a Whole

1.  Treemap: `{treemapify}` package
2.  Venn: `{VennDiagram}` package
3.  Donut & pie: `geom_bar()` and `coord_radial()` or `coord_polar()`
4.  Dendrogram: `{ggdendro}` package
5.  Circular packing: `{ggraph}` package

## Evolution

1.  Line: `geom_line()`
2.  Area & stacked area: `geom_area()`
3.  Stream: `{ggstream}` package
4.  Timeseries: `geom_point()`

## Map

1.  Choropleth, hexbin, cartogram, connection, and bubble maps: `{sf}` or `{tmap}` packages

## Flow

1.  Chord: `{chorddiag}` and `{circlize}` packages
2.  Network: `{igraph}` and `{ggraph}` packages
3.  Sankey: `{ggsankey}` and `{ggalluvial}` packages
4.  Arc: `{ggraph}` package
5.  [Edge bundling](https://www.data-to-viz.com/graph/edge_bundling.html): `{ggraph}` package

# Layer by Layer

One of the most powerful ideas in `ggplot2` is that plots are built **layer by layer**. That means you’re not limited to a single geom — you can stack multiple geoms on top of the same data, or even different data.

For example, you might want to:

-   Show individual points *and* a trend line
-   Create a time series with connected markers
-   Label a few important observations
-   Add summary statistics to a chart
-   Use an arrow or box to highlight a finding

The syntax is simple: just add more layers, as usual, with a `+` sign!

Now let’s look at an example that puts layering into action — combining points, trend lines, and labels with different data and aesthetics. It’s a great way to see how layering gives you precise control over what’s drawn.

```{r}
library(ggplot2)

ggplot(mpg, aes(x = displ, y = cty)) +
  # draw a scatter plot
  geom_point(
    aes(color = drv)
  ) + 
  # add a trend line
  geom_smooth(
    method = "lm"
  ) + 
  # add labels for extreme values
  geom_text(
    # pass a filtered dataset
    data = subset(mpg, cty > 30 | displ >= 6.5),
    aes(label = model),
    vjust = 1.4 # Move text below the points
  )

```

Each layer can:

-   Use the **same dataset** — like `geom_point()` and `geom_smooth()`
-   Use the **same aesthetics** — `x` and `y` are passed globally to all layers

*or:*

-   Use a **different dataset** — like `geom_text()`
-   Use **layer-specific aesthetics** — `color` is only mapped inside `geom_point()`; `label` only inside `geom_text()`

Don’t forget: layers expect the `mapping` argument first. If you want to customize the dataset, use `data =` after that.

Geoms are **additive** — think of it like building up a painting: first the background, then the foreground, then the annotations.

***That means: the order matters too!***

# Getting Things in Position

So far, we’ve seen how layers control *what* gets drawn — by choosing the right geom and mapping aesthetics appropriately.

But there’s another important setting that affects *where* elements appear: the `position` argument.

Many geoms — especially bars, areas, boxplots, and error bars — often display *multiple groups* within the same chart. The `position` argument controls how those groups are arranged: stacked, dodged side-by-side, overlaid, or normalized to proportions.

You can specify this argument in two main ways: \* Using strings like `"stack"` or `"jitter"` \* Using functions like `position_stack()` or `position_jitter()`

Both approaches work — but only the function versions allow you to *fine-tune* the behavior by passing arguments:

```{r}
# Position as a string
# Passing position as a string uses the default settings of the respective position_*() function — quick, but not customizable.

ggplot(mpg, aes(x = class, y = hwy)) + 
  geom_point(
    alpha = 0.5, size = 3, position = "jitter"
  )

# Position as a function
# Passing position as a function lets you customize its behavior — for example, adjusting the amount of jitter or fixing the random noise.

ggplot(mpg, aes(x = class, y = hwy)) + 
  geom_point(
    alpha = 0.5, size = 3, position = position_jitter(
      # limit horizontal spread
      width = 0.3,
      # ensure jittering only applies on x
      height = 0,
      # fix the random noise
      seed = 12345
    )
  )
```

“Hold up — what’s that seed thing doing?”

-   Run the code in the first tab multiple times.
-   Notice how the points jump around each time?
-   Now switch to the second tab and do the same.
-   The points stay put! Change the number to use a different pattern.

Tired of typing position = "jitter"?

You don’t have to! Instead of writing `geom_point(position = "jitter")` every time, you can use the shorthand wrapper `geom_jitter()` — it’s just `geom_point()` with `position = "jitter"` set as the default.

Just keep in mind: if you want to control how much jitter is applied or fix the randomness with a seed, you’ll still need to use `position_jitter()` directly.

Now let’s explore a few common cases where `position =` unlocks powerful layout options — and helps keep your chart clear and intentional.

# Flip the Perspective

Area charts stack by default — values for each series accumulate vertically. That’s great for cumulative trends, but not always what you want.

```{r}
df <- data.frame(
  time = rep(1:12, 2),
  value = c(1, 2, 5, 3, 7, 8, 6, 9, 5, 5, 8, 4,
            4, 5, 8, 4, 5, 7, 4, 5, 3, 4, 2, 3),
  group = rep(c("A", "B"), each = 12)
)

ggplot(df, aes(x = time, y = value)) + 
  geom_area(
    aes(color = group, fill = group),
    alpha = 0.5
    # and, by default: position = stack
  )
```

Note that `geom_area()`, similar to other geoms that use areas like bars or ribbons, **has two colors — an outline `color` and an inside `fill`**. In that case, the transparency, set via alpha = .5 is only applied to the fill.

You can change this default behavior by specifying `position = "identity"` to plot overlapping areas without stacking. Now, one series is plotted in front of the previous.

```{r}
ggplot(df, aes(x = time, y = value)) +
  geom_area(
    aes(color = group, fill = group),
    alpha = 0.5,
    position = "identity"
  )
```
And if you want to compare proportions rather than raw values, there’s a potentially lesser-known gem: `position = "fill"` — this normalizes the areas to show relative shares instead, without the need for pre-calculation!

```{r}
ggplot(df, aes(x = time, y = value)) +
  geom_area(aes(color = group, fill = group),
            alpha = 0.5,
            position = "fill")
```
Each of the different options changes the perspective — and the story — dramatically. You have to decide which supports your main message best.

# Group Up — Line Up — Dodge Right
Bar charts stack by default too — but that’s not always ideal. Often, we want to compare groups side by side.

The magic phrase? We want to _dodge_ the groups.

There are two different ways to change the behavior:

* `"dodge"` and `position_dodge()`
* `"dodge2"` or `position_dodge2()`

`position_dodge2()` gives you more control by offering additional arguments to control the `padding` or to `reverse` the order. By default, a padding of `0.1` is added to the bars.

```{r}
df <- data.frame(
  category = rep(c("A", "B", "C"), times = 2),
  value = c(2, 5, 0, 2, 3, 6),
  year = rep(c("2020", "2024"), each = 3)
)

# Stack (default)
ggplot(df, aes(x = category, y = value, fill = year)) + 
  geom_col(
    # position = "stack" # the default
  )

# Dodge
ggplot(df, aes(x = category, y = value, fill = year)) + 
  geom_col(
    position = "dodge" 
  )

# Dodge2
ggplot(df, aes(x = category, y = value, fill = year)) + 
  geom_col(
    position = "dodge2" 
  )
```
Dodging is not limited to bar or column charts — we can apply the same logic to other geoms as well.

Let's switch our column chart to a dot plot. And, to show off, label those points!

To avoid overplotting in category A, we are applying dodging to both `geom_point()` and `geom_text()` with the same width to keep them aligned.

Stacking, however, makes absolutely no sense in this case — don't do this at home!

```{r}
df <- data.frame(
  category = rep(c("A", "B", "C"), times = 2),
  value = c(2, 5, 0, 2, 3, 6),
  year = rep(c("2020", "2024"), each = 3)
)

# Default positions
ggplot(df, aes(x = category, y = value, fill = year)) +
  geom_point(
    # position = "identity", # the default
  ) +
  geom_text(
    aes(label = value),
    # position = "identity", # the default
    hjust = -1
  ) +
  geom_line(
    aes(group = category),
    color = "#d13760"
  )

# Dodged posiitons
ggplot(df, aes(x = category, y = value, fill = year)) +
  geom_point(
    position = position_dodge(width = 0.67)
  ) +
  geom_text(
    aes(label = value),
    position = position_dodge(width = 0.67),
    hjust = -1
  ) +
  geom_line(
    aes(group = category),
    color = "#d13760"
  )

# Stacked positions (wrong)
ggplot(df, aes(x = category, y = value, fill = year)) +
  geom_point(
    position = "stack"
  ) +
  geom_text(
    aes(label = value),
    position = "stack",
    hjust = -1
  ) +
  geom_line(
    aes(group = category),
    color = "#d13760"
  )
```
We have added lines to indicate the change in position — notice how they are completely off when using `"stack"`.

With a solid understanding of geoms and how to position layers effectively, combined with your earlier insights into the logic, tricks, and challenges of working with aesthetics, together, these give you a powerful foundation to craft nearly any type of chart with confidence.