# [6.1 Using geometry in tidycensus](https://walker-data.com/census-r/mapping-census-data-with-r.html#using-geometry-in-tidycensus)

tidycensus wraps several common geographic data functions in the **tigris** 
package to allow R users to return simple feature geometry pre-linked to 
downloaded demographic data with a single function call. The key argument to
accomplish this is `geometry = TRUE`, which is available in the core data download
functions in tidycensus, `get_acs()`, `get_decennial()`, and `get_estimates()`.

`geometry = TRUE` combines the automated data download functionality of 
tidycensus and tigris to allow R users to bypass this process entirely. 
The following example illustrates the use of the `geometry = TRUE` argument, 
fetching information on median household income for Census tracts in the 
District of Columbia. As discussed in the previous chapter, the option 
`tigris_use_cache = TRUE` is used to cache the downloaded geographic data on the 
user’s computer.

```{r}
library(tidycensus)
options(tigris_use_cache = T)

dc_income <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "DC",
  year = 2020,
  geometry = T
)

dc_income
```

As shown in the example call, the structure of the object returned by tidycensus 
resembles the object we’ve become familiar with to this point in the book. 
For example, median household income data are found in the `estimate` column with 
associated margins of error in the moe column, along with a variable ID, GEOID, 
and Census tract name. However, there are some notable differences. The geometry 
column contains polygon feature geometry for each Census tract, allowing for a 
linking of the estimates and margins of error with their corresponding locations 
in Washington, DC. Beyond that, the object is associated with coordinate system 
information - using the NAD 1983 geographic coordinate system in which Census 
geographic datasets are stored by default.

## 6.1.1 Basic mapping of sf objects with `plot()`
Such geographic information can be difficult to understand without visualization. 
As the returned object is a simple features object, both geometry and attributes 
can be visualized with `plot()`. Key here is specifying the name of the column to 
be plotted inside of brackets, which in this case is `"estimate"`.

```{r}
plot(dc_income["estimate"])
```

The `plot()` function returns a simple map showing income variation in Washington, DC. 
Wealthier areas, as represented with warmer colors, tend to be located in the 
northwestern part of the District. `NA` values are represented on the map in white. 
If desired, the map can be modified further with base plotting functions.

