# [6.2 Map-making with ggplot2 and geom_sf](https://walker-data.com/census-r/mapping-census-data-with-r.html#map-making-with-ggplot2-and-geom_sf)

`geom_sf()` in **ggplot2** can be used for quick plotting of sf objects using 
familiar **ggplot2** syntax. `geom_sf()` goes far beyond simple cartographic display. 
The full power of **ggplot2** is available to create highly customized maps and 
geographic data visualizations.

## 6.2.1 Choropleth mapping
One of the most common ways to visualize statistical information on a map is with 
`choropleth mapping`. Choropleth maps use shading to represent how underlying data 
values vary by feature in a spatial dataset. The income plot of Washington, DC 
shown earlier in this chapter is an example of a choropleth map.

In the example below, **tidycensus** is used to obtain linked ACS and spatial 
data on median age by state for the 50 US states plus the District of Columbia 
and Puerto Rico. For national maps, it is often preferable to generate insets of 
Alaska, Hawaii, and Puerto Rico so that they can all be viewed comparatively with 
the continental United States. We’ll use the `shift_geometry()` function in **tigris** 
to shift and rescale these areas for national mapping. The argument 
`resolution = "20m"` is necessary here for appropriate results, as it will omit 
the long archipelago of islands to the northwest of Hawaii.

```{r}
library(tidycensus)
library(tidyverse)
library(tigris)

us_median_age <- get_acs(
  geography = "state",
  variables = "B01002_001",
  year = 2019,
  survey = "acs1",
  geometry = T,
  resolution = "20m"
) %>% 
  shift_geometry()

plot(us_median_age$geometry)
```
The state polygons can be styled using **ggplot2** conventions and 
the `geom_sf()` function. With two lines of **ggplot2** code, a basic map of 
median age by state can be created with **ggplot2** defaults.

```{r}
ggplot(data = us_median_age, aes(fill = estimate)) + geom_sf()
```
The `geom_sf()` function in the above example interprets the geometry of the 
sf object (in this case, polygon) and visualizes the result as a filled 
choropleth map. In this case, the ACS estimate of median age is mapped to the 
default blue dark-to-light color ramp in **ggplot2**, highlighting the youngest 
states (such as Utah) with darker blues and the oldest states (such as Maine) 
with lighter blues.

## 6.2.2 Customizing ggplot2 maps
n many cases, map-makers using **ggplot2** will want to customize this graphic further. 
For example, a designer may want to modify the color palette and reverse it so 
that darker colors represent older areas. The map would also benefit from some 
additional information describing its content and data sources. 
These modifications can be specified in the same way a user would update a regular ggplot2 graphic. 
The `scale_fill_distiller()` function allows users to specify a [ColorBrewer](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) palette 
to use for the map, which includes a wide range of sequential, diverging, and 
qualitative color palettes. The `labs()` function can then be used to add a 
title, caption, and better legend label to the plot. 
Finally, **ggplot2** cartographers will often want to use the `theme_void()` function 
to remove the background and gridlines from the map.

```{r}
ggplot(data = us_median_age, aes(fill = estimate)) +
  geom_sf() +
  scale_fill_distiller(palette = "RdPu",
                       direction = 1) +
  labs(title = "  Median Age by State, 2019",
       caption = "Data source: 2019 1-year ACS, US Census Bureau",
       fill = "ACS estimate") +
  theme_void()
```
