[6.4 Cartographic workflows wtih non-Census data](https://walker-data.com/census-r/mapping-census-data-with-r.html#cartographic-workflows-with-non-census-data)

In many instances, an analyst may possess data that is available at a Census 
geography but is not available through the ACS or decennial Census. 
This means that the `geometry = TRUE` functionality in **tidycensus**, which 
automatically enriches data with geographic information, is not possible. 
In these cases, Census shapes obtained with **tigris** can be joined to tabular data 
and then visualized.

This section covers two such workflows. The first reproduces the popular 
red/blue election map common in presidential election cycles. 
The second focuses on mapping zip code tabulation areas, or ZCTAs, a geography 
that represents the spatial location of zip codes (postal codes) in the United States.

## 6.4.1 National election mapping with tigris shapes
While enumeration units like Census tracts and block groups will generally be 
used to map Census data, Census shapes representing legal entities are useful 
for a variety of cartographic purposes. A popular example is the political map, 
which shows the winner or poll results from an election in a region. 
We’ll use data from the Cook Political Report to generate a basic red state/blue 
state map of the 2020 US Presidential election results. 
This dataset was downloaded on June 5, 2021 and is available at 
`"data/us_vote_2020.csv"` in the book GitHub repository.

```{r}
library(tidyverse)
library(tigris)

vote2020 <- read_csv(
  "https://raw.githubusercontent.com/walkerke/census-with-r-book/master/data/us_vote_2020.csv"
)

names(vote2020)
```
The data include a wide variety of columns that can be visualized on a map. 
As discussed in the previous chapter, a comparative map of the United States 
can use the `shift_geometry()` function in the **tigris** package to shift and 
rescale Alaska and Hawaii. State geometries are available in **tigris** with the 
`states()` function, which should be used with the arguments `cb = TRUE` and 
`resolution = "20m"` to appropriately generalize the state geometries for national mapping.

To create the map, the geometry data obtained with tigris must be joined to the 
election data from the Cook Political Report. This is accomplished with the 
`left_join()` function from **dplyr**. **dplyr**’s `*_join()` family of functions are 
supported by simple features objects, and work in this context analogous to the
common “Join” operations in desktop GIS software. The join functions work by
matching values in one or more “key fields” between two tables and merging data
from those two tables into a single output table. The most common join functions
you’ll use for spatial data are `left_join()`, which retains all rows from the
first dataset and fills non-matching rows with `NA` values, and `inner_join()`,
which drops non-matching rows in the output dataset.

Let’s try this out by obtaining low-resolution state geometry with **tigris**,
shifting and rescaling with `shift_geometry()`, then merging the political data
to those shapes, matching the `NAME` column in `us_states` to the `state` 
column in `vote2020`.

```{r}
us_states <- states(cb = T, resolution = "20m") %>% 
  filter(NAME != "Puerto Rico") %>% 
  shift_geometry()

us_states_joined <- us_states %>% 
  left_join(vote2020, by = c("NAME" = "state"))
```
Before proceeding we’ll want to do some quality checks. 
In `left_join()`, values must match exactly between `NAME` and `state` to merge correctly
and this is not always guaranteed when using data from different sources. 
Let’s check to see if we have any problems:

```{r}
table(is.na(us_states_joined$state))
```
We’ve matched all the 50 states plus the District of Columbia correctly. 
In turn, the joined dataset has retained the shifted and rescaled geometry of 
the US states and now includes the election information from the tabular dataset
which can be used for mapping. To achieve this structure, specifying the
directionality of the join was critical. For spatial information to be retained
in a join, the spatial object _must_ be on the left-hand side of the join.
In our pipeline, we specified the `us_states` object first and used `left_join()`
to join the election information to the states object. If we had done this in
reverse, we would have lost the spatial class information necessary to make the map.

For a basic red state/blue state map using **ggplot2** and `geom_sf()`,
a manual color palette can be supplied to the `scale_fill_manual()` function,
filling state polygons based on the `called` column which represents the party
for whom the state was called.

```{r}
ggplot(us_states_joined, aes(fill = called)) +
  geom_sf(color = "white", lwd = 0.2) +
  scale_fill_manual(values = c("blue", "red")) +
  theme_void() +
  labs(fill = "Party",
       title = " 2020 US presidential election results by state",
       caption = "Note: Nebraska and Maine split electoral college votes by congressional district")
```
## 6.4.2 Understandign and working with ZCTAs
The most granular geography at which many agencies release data is at the zip code level. 
This is not an ideal geography for visualization, given that zip codes represent collections 
of US Postal Service routes (or sometimes even a single building, or Post Office box) 
that are not guaranteed to form coherent geographies. The US Census Bureau allows for an 
approximation of zip code mapping with Zip Code Tabulation Areas, or ZCTAs. 
ZCTAs are shapes built from Census blocks in which the most common zip code for addresses 
in each block determines how blocks are allocated to corresponding ZCTAs. 
While ZCTAs are not recommended for spatial analysis due to these irregularities, 
they can be useful for visualizing data distributions when no other granular geographies are available.

An example of this is [the Internal Revenue Service’s Statistics of Income (SOI) data](https://www.irs.gov/statistics/soi-tax-stats-individual-income-tax-statistics-2018-zip-code-data-soi), 
which includes a wide range of indicators derived from tax returns. The most detailed 
geography available is the zip code level in this dataset, meaning that within-county 
visualizations require using ZCTAs. Let’s read in the data for 2018 from the IRS website:

```{r}
irs_data <- read_csv(
  "https://www.irs.gov/pub/irs-soi/18zpallnoagi.csv"
)

ncol(irs_data)
```
The dataset contains 153 columns [which are identified in the linked codebook](https://www.irs.gov/pub/irs-soi/18zpdoc.docx). 
Geographies are identified by the `ZIPCODE` column, which shows aggregated data by
state (`ZIPCODE == "000000"`) and by zip code. We might be interested in understanding
the geography of self-employment income within a given region. 
We’ll retain the variables `N09400`, which represents the number of tax returns 
with self-employment tax, and `N1`, which represents the total number of returns.

```{r}
self_employment <- irs_data %>% 
  select(ZIPCODE, self_emp = N09400, total = N1)
```

From here, we’ll need to identify a region of zip codes for analysis. 
In **tigris**, the `zctas()` function allows us to fetch a Zip Code Tabulation Areas shapefile. 
Given that some ZCTA geography is irregular and sometimes stretches across multiple states, 
a shapefile for the entire United States must first be downloaded. 
It is recommended that shapefile caching with `options(tigris_use_cache = TRUE)` 
be used with ZCTAs to avoid long data download times.

The `starts_with` parameter in `zctas()` allows users to filter down 
ZCTAs based on a vector of prefixes, which can identify an area without using a 
spatial process. For example,  we can get ZCTA data near Boston, MA by using the
appropriate prefixes.

```{r}
library(mapview)
library(tigris)
options(tigris_use_cache = T)

boston_zctas <- zctas(
  cb = T,
  starts_with = c("021", "022", "024"),
  year = 2018
)
```

Next we can use `mapview()` to inspect the results:

```{r}
mapview(boston_zctas)
```
The ZCTA prefixes `021`, `022`, and `024` cover much of the Boston metropolitan
area; "holes" in the region represent areas like Boston Common which are not
covered by ZCTAs. Let's take a quick look at its attributes:

```{r}
names(boston_zctas)
```

Either the `ZCTA4CE10` column or the `GEOID10` column can be matched to the 
appropriate zip code information in the IRS dataset for visualization. 
The code below joins the IRS data to the spatial dataset and computes a new 
column representing the percentage of returns with self-employment income.

```{r}
boston_se_data <- boston_zctas %>% 
  left_join(self_employment, by = c("GEOID10" = "ZIPCODE")) %>% 
  mutate(pct_self_emp = 100 * (self_emp / total)) %>% 
  select(GEOID10, self_emp, pct_self_emp)
  
```
There are a variety of ways to visualize this information. 
One such method is a choropleth map, which you’ve learned about earlier this chapter:

```{r}
library(tmap)
tm_shape(boston_se_data, projection = 26918) +
  tm_fill(col = "pct_self_emp",
          palette = "Purples",
          title = "% self-employed, \n2018 IRS SOI data")
```
The choropleth map shows that self-employment filings are more common in 
suburban Boston ZCTAs than nearer to the urban core, generally speaking. 
However, we might also be interested in understanding where most self-employment 
income filings are located rather than their share relative to the total number 
of returns filed. This requires visualizing the `self_emp` column directly. 
As discussed earlier in this chapter, a graduated symbol map with `tm_bubbles()` 
is preferable to a choropleth map for this purpose.

```{r}
tm_shape(boston_se_data) +
  tm_polygons() +
  tm_bubbles(size = "self_emp",
             alpha = 0.5,
             col = "navy",
             title.size = "Self-employed filers, \n2018 IRS SOI data")
```









