[6.5 Interactive Mapping](https://walker-data.com/census-r/mapping-census-data-with-r.html#interactive-mapping)

Modern web technologies - and the integration of those technologies into R with 
the `htmlwidgets` package, as discussed in Section 4.7.4 - 
make the creation of interactive, explorable Census data maps straightforward.

## 6.5.1 Interactive mapping with Leaflet
With over 31,000 GitHub stars as of July 2021, the Leaflet JavaScript library 
is one of the most popular frameworks worldwide for interactive mapping. 
The RStudio team brought the Leaflet to R with the **leaflet** R package, 
which now powers several approaches to interactive mapping in R. 
The following examples cover how to visualize Census data on an interactive 
Leaflet map using approaches from **mapview**, **tmap**, 
and the core **leaflet** package.

Let’s start by getting some illustrative data on the percentage of the 
population aged 25 and up with a bachelor’s degree or higher from the 
2016-2020 ACS. We’ll look at this information by Census tract in 
Dallas County, Texas.

```{r}
library(tidycensus)

dallas_bachelors <- get_acs(
  geography = "tract",
  variables = "DP02_0068P",
  year = 2020,
  state = "TX",
  county = "Dallas",
  geometry = T
)
```
In Chapter 5, you learned how to quickly visualize geographic data obtained 
with tigris on an interactive map by using the `mapview()` function in the 
**mapview** package. The `mapview()` function also includes a parameter `zcol` that 
takes a column in the dataset as an argument, and visualizes that column with
an interactive choropleth map.


```{r}
library(mapview)
mapview(dallas_bachelors, zcol = "estimate")
```

Conversion of **tmap** maps to interactive Leaflet maps is also straightforward with 
the command `tmap_mode("view")`. After entering this command, all subsequent tmap 
maps in your R session will be rendered as interactive Leaflet maps using the 
same tmap syntax you’d use to make static maps.

```{r}
library(tmap)
tmap_mode("view")

tm_shape(dallas_bachelors) +
  tm_fill(col = "estimate", palette = "magma", alpha = 0.5)
```

To switch back to static plotting mode, run the command `tmap_mode("plot")`.

For more fine-grained control over your Leaflet maps, the core **leaflet** package 
can be used. Below, we’ll reproduce the **mapview/tmap** examples using the leaflet 
package’s native syntax. First, a color palette will be defined using the 
`colorNumeric()` function. This function itself creates a function we’re calling 
`pal()`, which translates data values to color values for a given color palette. 
Our chosen color palette in this example is the viridis magma palette.

```{r}
library(leaflet)

pal <- colorNumeric(
  palette = "magma",
  domain = dallas_bachelors$estimate
)

pal(c(10, 20, 30, 40, 50))
```

The map itself is built with a magrittr pipeline and the following steps:

*The `leaflet()` function initializes the map. 
A `data` object can be specified here or in a function that comes later in the pipeline.

*`addProviderTiles()` helps you add a basemap to the map that will be shown 
beneath your data as a reference. Several providers are built-in to the Leaflet 
package, including the popular [Stamen](https://stamen.com/) reference maps. 
If you are only interested in a basic basemap, the `addTiles()` function returns 
the standard [OpenStreetMap](https://www.openstreetmap.org/#map=5/38.007/-95.844) basemap. 
Use the built-in `providers` object to try out 
different basemaps; it is good practice for choropleth mapping to use a 
greyscale or muted basemap.

*`addPolygons()` adds the tract polygons to the map and styles them. 
In the code below, we are using a series of options to specify the input data; 
to color the polygons relative to the defined color palette; 
and to adjust the smoothing between polygon borders, the opacity, 
and the line weight. The label argument will create a hover tooltip on the map 
for additional information about the polygons.

*`addLegend()` then creates a legend for the map, providing critical information 
about how the colors on the map relate to the data values.

```{r}
leaflet() %>% 
  addProviderTiles(providers$Stadia.StamenTonerLite) %>% 
  addPolygons(data = dallas_bachelors,
              color = ~pal(estimate),
              weight = 0.5,
              smoothFactor = 0.2,
              fillOpacity = 0.5,
              label = ~estimate) %>% 
  addLegend(
    position = "bottomright",
    pal = pal,
    values = dallas_bachelors$estimate,
    title = "% wtih bachelor's<br/>degree"
  )
```
This example only scratches the surface of what the **leaflet** R package can
accomplish; I'd encourage you [review the documentation for more examples](https://rstudio.github.io/leaflet/).

## 6.5.2 Alternative approaches to interactive mapping
Like most interactive mapping platforms, Leaflet uses [_tiled mapping_](https://wiki.openstreetmap.org/wiki/Slippy_Map) in the 
Web Mercator coordinate reference system. Web Mercator works well for tiled web 
maps that need to fit within rectangular computer screens, and preserves angles 
at large scales (zoomed-in areas) which is useful for local navigation. 
However, it grossly distorts the area of geographic features near the poles, 
making it inappropriate for small-scale thematic mapping of the world or world regions.

Let’s illustrate this by mapping median home value by state from the 1-year ACS 
using **leaflet**. We’ll first acquire the data with geometry using tidycensus, 
setting the output resolution to “20m” to get low-resolution boundaries to speed 
up our interactive mapping.

```{r}
us_value <- get_acs(
  geography = "state",
  variables = "B25077_001",
  year = 2019,
  survey = "acs1",
  geometry = T,
  resolution = "20m"
)
```
The acquired ACS data for the US can be mapped using the same techniques as 
with the educational attainment map for Dallas County.

```{r}
us_pal <- colorNumeric(
  palette = "plasma",
  domain = us_value$estimate
)

leaflet() %>% 
  addProviderTiles(providers$Stadia.StamenTonerLite) %>% 
  addPolygons(data = us_value,
              color = ~us_pal(estimate),
              weight = 0.5,
              smoothFactor = 0.2,
              fillOpacity = 0.5,
              label = ~estimate) %>% 
  addLegend(
    position = "bottomright",
    pal = us_pal,
    values = us_value$estimate,
    title = "Median home value"
  )
```
The disadvantages of Web Mercator - as well as this general approach to mapping 
the United States - are on full display. Alaska’s area is grossly distorted 
relative to the rest of the United States. It is also difficult on this map to 
compare Alaska and Hawaii to the continental US - which is particularly important 
in this example as Hawaii’s median home value is the highest in the entire country. 
The solution proposed elsewhere in this book is to use `tigris::shift_geometry()` 
which adopts appropriate coordinate reference systems for Alaska, Hawaii, 
and the continental US and arranges them in a better comparative fashion on the map. 
However, this approach risks losing the interactivity of a Leaflet map.

A compromise solution can involve other R packages that allow for interactive mapping. 
An excellent option is the **ggiraph** package, which like the plotly package can 
convert ggplot2 graphics into interactive plots. In the example below, 
interactivity is added to a ggplot2 plot with **ggiraph**, allowing for panning and 
zooming with a hover tooltip on a shifted and rescaled map of the US.

```{r}
library(ggiraph)
library(scales)
library(dplyr)
library(tigris)
library(ggplot2)

us_value_shifted <- us_value %>% 
  shift_geometry(position = "outside") %>% 
  mutate(tooltip = paste(NAME, estimate, sep = ": "))

gg <- ggplot(us_value_shifted, aes(fill = estimate)) +
  geom_sf_interactive(aes(tooltip = tooltip, data_id = NAME),
                      size = 0.1) +
  scale_fill_viridis_c(option = "plasma", labels = label_dollar()) +
  labs(title = "Median housing value by State, 2019",
       caption = "Data source: 2019 1-year ACS, US Census Bureau",
       fill = "ACS estimate") +
  theme_void()

girafe(ggobj = gg) %>% 
  girafe_options(opts_hover(css = "fill:cyan;"),
                 opts_zoom(max = 10))
```



