# Change Your Style

Themes are a **different kind of component**: they act as the “finishing layer” of your visualization and focus entirely on **styling and presentation**. Typefaces, font styles, legends, backgrounds, grid lines, margins — all of these can be controlled with themes to make your charts look clean and polished without touching the underlying data.

Themes let you give them a _cohesive look_ to make it publication-ready — simply swap a complete theme and or nudge a few elements to add a custom and personal touch.

# Themes at a Glance
The fastest way to style your plot is by applying a _complete theme_.

A theme is simply a pre-built set of settings that control everything—from background colors and grid lines to text and spacing. With just one extra line of code, you get a clean, consistent style to build on.

Here’s an example using the minimal theme:

```{r}
library(ggplot2)
penguins <- read.csv("https://www.ggplot2-uncharted.com/data/palmerpenguins.csv")

ggplot(penguins) +
  aes(
    x = flipper_length_mm,
    y = bill_length_mm,
    color = species
  ) +
  geom_point() +
  ggtitle("Bill ratios of Pygoscelis penguins") +
  theme_minimal()

```
Note: By default, every ggplot starts with the complete theme `theme_grey()`, which adds the classic “corporate grey” look. That’s a solid baseline — but it can feel a bit… well, _ggplot-y_.

# World of Themes
The `ggplot2` package comes with a set of complete themes to cover common needs.

Beyond that, the community has created packages with additional ready-to-use themes, such as [ggthemes](https://jrnold.github.io/ggthemes/), [hrbrthemes](https://github.com/hrbrmstr/hrbrthemes), and [tvthemes](https://ryo-n7.github.io/tvthemes/).

Some replicate the visual styles of well-known companies, newspapers, or software, while others take a more playful approach—referencing popular TV shows and other bits of pop culture.

# Hello, Base
When applying a complete theme, two arguments are especially handy: `base_size` and `base_family`. 
They let you adjust the base settings in one go.

* `base_size` sets the overall scale of all theme elements
* `base_family` defines the typeface for all text elements

Note: `base_line_size` and `base_rect_size` also exist, but as these scale with `base_size`,
you'll rarely need to touch them.

This way, you don’t need to repeat those settings for each element to change the overall readability and character of a plot: adjusting text size and typeface alone can already make a chart feel more professional, playful, or simply easier to read in context.

```{r}
ggplot(penguins) +
  aes(x = bill_length_mm,
      y = bill_depth_mm,
      color = species) +
  geom_point() +
  ggtitle("Bill ratios of Pygoscelis penguins") +
  theme_grey(
    base_size = 15,
    base_family = "monospace"
  )
```

## Which Fonts Can I Use in `base_family`
While you can only specify the **universal font families** in our sandboxes — `"sans"` (default), `"serif"`, and `"monospace"` — you’re not limited to these in your own setup. Any installed font can be applied to your ggplot themes using `base_family`.

For example, the font we use for the body text of this course, `Domine`, can be installed locally and then referenced in your themes by its family name:

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar() +
  theme_minimal(base_family = "Domine")
```

We will talk about using non-default typefaces a lot more in upcoming lessons. For a detailed guide on using custom fonts in R and applying them in your plots, see the [Custom Themes lesson](https://www.ggplot2-uncharted.com/module2/custom-themes).

# Going Down the Rabbit Hole
Adjusting the base settings of a complete theme won’t affect colors or the layout of individual elements. That’s where the `theme()` function comes in: in the next lesson, you’ll learn how to fine-tune your plots in detail — from changing the typeface of a specific text element, to tweaking grid lines and margins, to moving the legend exactly where you want it.

Still, complete themes, whether built into `ggplot2` or provided by extension packages, already give you a strong foundation. And if you prefer to keep things simple, you can stick with those plus a few base settings for clean, polished plots without much extra effort.

