# [Pivoting](https://tidyr.tidyverse.org/articles/pivot.html)

## Introduction 

This vignette describes the use of new `pivot_longer()` and `pivot_wider()` functions. Their goal is to improve the usability of `gather()` and `spread()`, and incorporate state-of-the-art-features found in other packages.

For some time, it's been obvious that there is something fundamentally wrong with the design of `spread()` and `gather()`. Many people don't find the names intuitive and find it hard to remember which direction corresponds to spreading and which to gathering. It also seems surprisingly hard to remember the arguments to these functions, meaning that many people (including me!) have to consult the documentation every time.

There are two important new features inspired by other R packages that have been advancing reshaping in R:

* `pivot_longer()` can work with multiple value variables that may have different types, inspired by the enhanced `melt()` and `dcast()` functions provided by the [data.table](https://github.com/Rdatatable/data.table/wiki) package by Matt Dowle and Arun Srinivasan.

* `pivot_longer()` and `pivot_wider()` can take a data frame that specifies precisely how metadata stored in column names becomes data variables (and vice versa), inspired by the [cdata](https://winvector.github.io/cdata/) package by John Mount and Nina Zumel.

In this vignette you'll learn the key ideas behind `pivot_longer()` and `pivot_wider()` as you see them used to solve a variety of data reshaping challenges ranging from simple to complex.

To begin we'll load some needed packages. In real analysis code, I'd imagine you'd do with the `library(tidyverse)`, but I can't do that here since this vignette is embedded in a package.

```{r}
library(tidyr)
library(dplyr)
library(readr)
```

## Longer
`pivot_longer()` makes datasets **longer** by increasing the number of rows and decreasing the number of columns. I don't believe it makes sense to describe a dataset as being in "long form". Length is a relative term, and you can only say (e.g.) that dataset A is longer than dataset B.

`pivot_longer()` is commonly needed to tidy wild-caught datasets as they often optimize for ease of data entry or ease of comparison rather than ease of analysis. The following sections show how to use `pivot_longer()` for a wide range of realistic datasets.

### String data in column names
The `relig_income` dataset stores counts based on a survey which (among other things) 

```{r} relig_income```

This dataset contains three variables:

* `religion`, stored in the rows,
* `income` spread across the column names, and
* `count` stored in the cell values.

To tidy it we use `pivot_longer()`:

```{r}
relig_income %>%
    pivot_longer(
        cols = !religion,
        names_to = "income",
        values_to = "count"
    )
```

* The first argument is the dataset to reshape, `relig_income`.

* `cols` describes which columns need to be reshaped. In this case, it's every column apart from `religion`.

* `names_to` gives the name of the variable that will be created from the data stored in the column names, i.e. `income`.

* `values_to` gives the name of the variable that will be created from the data stored in the cell value, i.e. `count`.

Neither the `names_to` nor the `values_to` column exists in `relig_income`, so we provide them as strings surrounded by quotes.

### Numeric data in column names.

The `billboard` dataset records the billboard rank of songs in the year 2000. It has a form similar to the `relig_income` data, but the data encoded in the column names is really an number, not a string.

```{r} billboard```

We can start with the 