# [7.4 Distance and proximity analysis](https://walker-data.com/census-r/spatial-analysis-with-us-census-data.html#distance-and-proximity-analysis)

A common use case for spatially-referenced demographic data is the analysis of _accessibility_. This might include studying the relative accessibility of different demographic groups to resources within a given region, or analyzing the characteristics of potential customers who live within a given distance of a store. Conceptually, there are a variety of ways to measure accessibility. The most straightforward method, computationally, is using straight-line (Euclidean) distances over geographic data in a projected coordinate system. A more computationally complex - but potentially more accurate - method involves the use of transportation networks to model accessibility, where proximity is measured not based on distance from a given location but instead based on travel times for a given transit mode, such as walking, cycling, or driving. This section will illustrate both types of approaches. Let’s consider the topic of accessibility to Level I and Level II trauma hospitals by Census tract in the state of Iowa. 2019 Census tract boundaries are acquired from **tigris**, and we use `st_read()` to read in a shapefile of hospital locations acquired from the US Department of Homeland Security.

```{r}
library(tigris)
library(sf)
library(tidyverse)
library(jsonlite)

options(tigris_use_cache = T)

# CRS: NAD83 / Iowa North
ia_tracts <- tracts("IA", cb = T, year = 2019) %>% 
  st_transform(26975)

hospital_url <- "https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Hospitals_gdb/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

trauma <- st_read(hospital_url) %>%
  filter(str_detect(TRAUMA, "LEVEL I\\b|LEVEL II\\b|RTH|RTC")) %>%
  st_transform(26975) %>%
  distinct(ID, .keep_all = TRUE)

```

## 7.4.1 Calculating distances

To determine accessibility of Iowa Census tracts to Level I or II trauma centers, we need to identify not only those hospitals that are located in Iowa, but also those in other states near to the Iowa border, such as in Omaha, Nebraska and Rock Island, Illinois. We can accomplish this by applying a distance threshold in `st_filter()`. In this example, we use the spatial predicate `st_is_within_distance`, and set a 100km distance threshold with the `dist = 100000` argument (specified in meters, the base measurement unit of our coordinate system used).

```{r}
ia_trauma <- trauma %>% 
  st_filter(ia_tracts,
            .predicate = st_is_within_distance,
            dist = 100000)

ggplot() + 
  geom_sf(data = ia_tracts, color = "NA", fill = "grey50") + 
  geom_sf(data = ia_trauma, color = "red") + 
  theme_void()

```
As illustrated in the visualization, the `st_filter()` operation has retained Level I and II trauma centers _within_ the state of Iowa, but also within the 100km threshold beyond the state’s borders.

With the Census tract and hospital data in hand, we can calculate distances from Census tracts to trauma centers by using the `st_distance()` function in the **sf** package. `st_distance(x, y)` by default returns the dense matrix of distances computed from the geometries in `x` to the geometries in `y`. In this example, we will calculate the distances from the _centroids_ of Iowa Census tracts (reflecting the center points of each tract geometry) to each trauma center.

```{r}
dist <- ia_tracts %>%
  st_centroid() %>%
  st_distance(ia_trauma) 

dist[1:5, 1:5]
```

A glimpse at the matrix shows distances (in meters) between the first five Census tracts in the dataset and the first five hospitals. When considering _accessibility_, we may be interested in the distance to the _nearest_ hospital to each Census tract. The code below extracts the minimum distance from the matrix for each row, converts to a vector, and divides each value by 1000 to convert values to kilometers. A quick histogram visualizes the distribution of minimum distances.

```{r}
min_dist <- dist %>% 
  apply(1, min) %>% 
  as.vector() %>% 
  magrittr::divide_by(1000)

hist(min_dist)
```


* The `apply()` function from base R is used to iterate over rows of the matrix. Matrices are a data structure not handled by the `map_*()` family of functions in the tidyverse, so base R methods must be used. In the example pipeline, the `apply()` function inherits the dist matrix object as its first argument. The second argument, `1`, refers to the margin of the matrix that `apply()` will iterate over; `1` references rows (which we want), whereas `2` would be used for columns. `min` then is the function to be applied to each row, giving us the minimum distance to a hospital for each Census tract.

* The `divide_by()` function in the **magrittr** package is a convenience arithmetic function to be used in analytic pipelines as R’s arithmetic operators (e.g. `/` for division) won’t work in this way. In this example, it divides all the values by 1000 to convert meters to kilometers.

While many tracts are within 10km of a trauma center, around 16 percent of Iowa Census tracts in 2019 are beyond 100km from a Level I or II trauma center, suggesting significant accessibility issues for these areas.

## 7.4.2 Calculating travel times

An alternative way to model accessibility to hospitals is through _travel times_ rather than distance, as the way that people experience access to locations is through time expended given a transportation network. While network-based accessibility may be a more accurate representation of people’s lived experiences, it is more computationally complex and requires additional tools. To perform spatial network analyses, R users will either need to obtain network data (like roadways) and use appropriate tools that can model the network; set up a routing engine that R can connect to; or connect to a hosted routing engine via a web API. In this example, we’ll use the **mapboxapi** R package to perform network analysis using [Mapbox’s travel-time Matrix API](https://docs.mapbox.com/api/navigation/matrix/).

The function `mb_matrix()` in **mapboxapi** works much like `st_distance()` in that it only requires arguments for origins and destinations, and will return the dense matrix of travel times by default. In turn, much of the computational complexity of routing is abstracted away by the function. However, as routes will be computed across the state of Iowa and API usage is subject to rate-limitations, the function can take several minutes to compute for larger matrices like this one.

If you are using **mapboxapi** for the first time, visit mapbox.com, register for an account, and obtain an access token. The function `mb_access_token()` installs this token in your `.Renviron` for future use.

```{r}
library(mapboxapi)
# mb_access_token("pk.eybcasq...", install = T)

times <- mb_matrix(ia_tracts, ia_trauma)

times[1:5, 1:5]
```

A glimpse at the travel-time matrix shows a similar format to the distance matrix, but with travel times in minutes used instead of meters. As with the distance-based example, we can determine the minimum travel time from each tract to a Level I or Level II trauma center. In this instance, we will visualize the result on a map.

```{r}
min_time <- apply(times, 1, min)

ia_tracts$time <- min_time

ggplot(ia_tracts, aes(fill = time)) + 
  geom_sf(color = NA) + 
  scale_fill_viridis_c(option = "magma") + 
  theme_void() + 
  labs(fill = "Time (minutes)",
       title = "Travel time to nearest Level I or Level II trauma hospital",
       subtitle = "Census tracts in Iowa",
       caption = "Data sources: US Census Bureau, US DHS, Mapbox")
```

The map illustrates considerable accessibility gaps to trauma centers across the state. Whereas urban residents typically live within 20 minutes of a trauma center, travel times in rural Iowa can exceed two hours.

An advantage to using a package like **mapboxapi** for routing and travel times is that users can connect directly to a hosted routing engine using an API. Due to rate limitations, however, web APIs are likely inadequate for more advanced users who need to compute travel times at scale. There are several R packages that can connect to user-hosted routing engines which may be better-suited to such tasks. These packages include [osrm](https://github.com/rCarto/osrm) for the Open Source Routing Machine; [opentripplanner](https://docs.ropensci.org/opentripplanner/) for OpenTripPlanner; and [r5r](https://ipeagit.github.io/r5r/) for R5.

## 7.4.3 Catchment areas with buffers and isochrones

The above example considers a broader accessibility analysis across the state of Iowa. In many cases, however, you’ll want to analyze accessibility in a more local way. A common use case might involve a study of the demographic characteristics of a hospital _catchment_ area, defined as the area around a hospital from which patients will likely come.

As with the matrix-based accessibility approach outlined above, catchment area-based proximity can be modeled with either Euclidean distances or network travel times as well. Let’s consider the example of [Iowa Methodist Medical Center in Des Moines](https://www.unitypoint.org/desmoines/iowa-methodist-medical-center.aspx), one of two Level I trauma centers in the state of Iowa.

The example below illustrates the distance-based approach using a _buffer_, implemented with the `st_buffer()` function in **sf**. A buffer is a common GIS operation that represents the area within a given distance of a location. The code below creates a 5km buffer around Iowa Methodist Medical Center by using the argument `dist = 5000`.

```{r}
iowa_methodist <- filter(ia_trauma, ID == "0009850308")

buf5km <- st_buffer(iowa_methodist, dist = 5000)
```
An alternative option is to create network-based _isochrones_, which are polygons that represent the accessible area around a given location within a given travel time for a given travel mode. Isochrones are implemented in the **mapboxapi** package with the `mb_isochrone()` function. Mapbox isochrones default to typical driving conditions around a location; this can be adjusted with the `depart_at` parameter for historical traffic and the argument `profile = "driving-traffic"`. The example below draws a 10-minute driving isochrone around Iowa Methodist for a Tuesday during evening rush hour.

```{r}
iso10min <- mb_isochrone(
  iowa_methodist, 
  time = 10, 
  profile = "driving-traffic",
  depart_at = "2022-04-05T17:00"
  )
```

We can visualize the comparative extents of these two methods in Des Moines. Run the code on your own computer to get a synced interactive map showing the two methods. The `makeAwesomeIcon()` function in **leaflet** creates a custom icon appropriate for a medical facility; [many other icons are available for common points of interest](http://rstudio.github.io/leaflet/markers.html#awesome-icons).

```{r}
library(leaflet)
library(leafsync)

hospital_icon <- makeAwesomeIcon(icon = "ios-medical", 
                                 markerColor = "red",
                                 library = "ion")

# The Leaflet package requires data be in CRS 4326
map1 <- leaflet() %>% 
  addTiles() %>%
  addPolygons(data = st_transform(buf5km, 4326)) %>% 
  addAwesomeMarkers(data = st_transform(iowa_methodist, 4326),
                    icon = hospital_icon)

map2 <- leaflet() %>% 
  addTiles() %>%
  addPolygons(data = iso10min) %>% 
  addAwesomeMarkers(data = st_transform(iowa_methodist, 4326),
                    icon = hospital_icon)

sync(map1, map2)
```

The comparative maps illustrate the differences between the two methods quite clearly. Many areas of equal distance to the hospital do not have the same level of access; this is particularly true of areas to the south of the Raccoon/Des Moines River. Conversely, due to the location of highways, there are some areas outside the 5km buffer area that can reach the hospital within 10 minutes.

## 7.4.4 Computing demographic estimates for zones with areal interpolation

Common to both methods, however, is a mis-alignment between their geometries and those of any Census geographies we may use to infer catchment area demographics. As opposed to the spatial overlay analysis matching Census tracts to metropolitan areas earlier in this chapter, Census tracts or block groups on the edge of the catchment area will only be partially included in the catchment. Areal interpolation methods like those introduced in Section 7.3 can be used here to estimate the demographics of both the buffer zone and isochrone.

Let’s produce interpolated estimates of the percentage of population in poverty for both catchment area definitions. This will require obtaining block group-level poverty information from the ACS for Polk County, Iowa, which encompasses both the buffer and the isochrone. The variables requested from the ACS include the number of family households with incomes below the poverty line along with total number of family households to serve as a denominator.

```{r}
library(tidycensus)

polk_poverty <- get_acs(
  geography = "block group",
  variables = c(poverty_denom = "B17010_001",
                poverty_num = "B17010_002"),
  state = "IA",
  county = "Polk",
  geometry = T,
  output = "wide",
  year = 2020
) %>% 
  select(poverty_denomE, poverty_numE) %>% 
  st_transform(26975)
```

We can then use population-weighted areal interpolation with `interpolate_pw()` function in **tidycensus** to estimate family poverty in both the buffer zone and the isochrone. Block weights for Polk County are obtained with **tigris**, and both the numerator and denominator columns are interpolated.

```{r}
library(glue)

polk_blocks <- blocks(
  state = "IA",
  county = "Polk",
  year = 2020
)

buffer_pov <- interpolate_pw(
  from = polk_poverty,
  to = buf5km,
  extensive = T,
  weights = polk_blocks,
  weight_column = "POP20",
  crs = 26975
) %>% 
  mutate(pct_poverty = 100 * (poverty_numE / poverty_denomE))

iso_pov <- interpolate_pw(
  from = polk_poverty, 
  to = iso10min,
  extensive = TRUE,
  weights = polk_blocks,
  weight_column = "POP20",
  crs = 26975
) %>%
  mutate(pct_poverty = 100 * (poverty_numE / poverty_denomE))
```

The two methods return slightly different results, illustrating how the definition of catchment area impacts downstream analyses.





