# [4.1 Basic Census visualization with ggplot2](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html)

A critical part of the Census data analysis process is __data visualization__, 
where an analyst examines patterns and trends found in their data graphically. 
In many cases, the exploratory analyses outlined in the previous two chapters 
would be augmented significantly with accompanying graphics.

To get started, we’ll return to a dataset used in Section [2.4.2](https://walker-data.com/census-r/an-introduction-to-tidycensus.html#renaming-variable-ids), 
which includes data on median household income and median age by county in the 
state of Georgia from the 2016-2020 ACS. We are requesting the data in wide format, 
which will spread the estimate and margin of error information across the columns.

```{r}
library(tidycensus)

ga_wide <- get_acs(
  geography = "county",
  state = "Georgia",
  variables = c(medinc = "B19013_001",
                medage = "B01002_001"),
  output = "wide",
  year = 2020
)

```

## 4.1.1 Getting started with ggplot2

*ggplot2* visualizations are initialized with the `ggplot()` function, 
to which a user commonly supplies a dataset and an aesthetic, 
defined with the `aes()` function. Within the `aes()` function, 
a user can specify a series of mappings onto either the data axes or 
other characteristics of the plot, such as element fill or color.

After initializing the ggplot object, users can layer plot elements onto the plot object. 
Essential to the plot is a `geom`, which specifies one of many chart types available in *ggplot2*. 
For example, `geom_bar()` will create a bar chart, `geom_line()` a line chart, 
`geom_point()` a point plot, and so forth. 
Layers are linked to the ggplot object by using the `+` operator.

One of the first exploratory graphics an analyst will want to produce when 
examining a new dataset is a __histogram__, which characterizes the distribution of 
values in a column through varying lengths of bars. 
This first example uses *ggplot2* and its `geom_histogram()` function to generate 
such a histogram of median household income by county in Georgia. 
The optional call to `options(scipen = 999)` instructs R to avoid using scientific 
notation in its output, including on the *ggplot2* tick labels.

```{r}
library(tidyverse)

options(scipen = 999)

ggplot(ga_wide, aes(x = medincE)) + geom_histogram()
```

The histogram shows that the modal median household income of Georgia counties 
is around $40,000 per year, with a longer tail of wealthier counties on the 
right-hand side of the plot. In the histogram, counties are organized into “bins”, 
which are groups of equal width along the X-axis. The Y-axis then represents the 
number of counties that fall within each bin. By default, *ggplot2* organizes the 
data into 30 bins; this option can be changed with the bins parameter. 
For example, we can re-make the visualization with half the previous number of 
bins by including the argument `bins = 15` in our call to `geom_histogram()`.

```{r}
ggplot(ga_wide, aes(x = medincE)) +
  geom_histogram(bins = 15)
```

Histograms are not the only options for visualizing univariate data distributions. 
A popular alternative is the __box-and-whisker plot__, which is implemented in 
ggplot2 with `geom_boxplot()`. In this example, the column `medincE` is passed to 
the `y` parameter instead of `x`; this creates a vertical rather than horizontal box plot.

```{r}
ggplot(ga_wide, aes(y = medincE)) +
  geom_boxplot()
```

The graphic visualizes the distribution of median household incomes by county in 
Georgia with a number of different components. The central __box__ covers the 
interquartile range (the IQR, representing the 25th to 75th percentile of values 
in the distribution) with a central line representing the value of the 
distribution’s median. The __whiskers__ then extend to either the minimum and 
maximum values of the distribution or 1.5 times the IQR. In this example, 
the lower whisker extends to the minimum value, and the upper whisker extends to 
1.5 times the IQR. Values beyond the whiskers are represented as __outliers__ on the 
plot with points.

## 4.1.2 Visualizing multivariate relationships with scatter plots

As part of the exploratory data analysis process, analysts will often want to 
visualize __interrelationships between Census variables__ along with the univariate 
data distributions discussed above. For two numeric variables, a common 
exploratory chart is a __scatter plot__, which maps values in one column to the 
X-axis and values in another column to the Y-axis. The resulting plot then gives 
the analyst a sense of the nature of the relationship between the two variables.

Scatter plots are implemented in *ggplot2* with the `geom_point()` function, 
which plots points on a chart relative to X and Y values for observations in a dataset. 
This requires specification of two columns in the call to `aes()` as opposed to 
the single column used in the univariate distribution visualization examples. 
The example that follows generates a scatter plot to visualize the relationship 
between county median age and county median household income in Georgia.

```{r}
ggplot(ga_wide, aes(x = medageE, y = medincE)) +
  geom_point()
```

The graphic shows a cloud of points that in some cases can suggest the nature of 
the correlation between the two columns. In this example, however, the 
correlation is not immediately clear from the distribution of points. 
Fortunately, *ggplot2* includes the ability to “layer on” additional chart 
elements to help clarify the nature of the relationship between the two columns. 
The `geom_smooth()` function draws a fitted line representing the relationship 
between the two columns on the plot. The argument `method = "lm"` draws a straight 
line based on a linear model fit; smoothed relationships can be visualized as 
well with `method = "loess"`.

```{r}
ggplot(ga_wide, aes(x = medageE, y = medincE)) +
  geom_point() +
  geom_smooth(method = "lm")
```

The regression line suggests a modest negative relationship between the two columns, 
showing that county median household income in Georgia tends to decline slightly as median age increases.



















