# [4.2 Customizing ggplot2 visualizations](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#customizing-ggplot2-visualizations)

The attractive defaults of *ggplot2* visualizations allow for the creation of 
legible graphics with little to no customization. This helps greatly with 
exploratory data analysis tasks where the primary audience is the analyst 
exploring the dataset. Analysts planning to present their work to an external 
audience, however, will want to customize the appearance of their plots beyond 
the defaults to maximize interpretability. This section covers how to take a 
Census data visualization that is relatively illegible by default and polish 
it up for eventual presentation and export from R.

In this example, we will create a visualization that illustrates the percent of 
commuters that take public transportation to work for the largest metropolitan 
areas in the United States. The data come from the 2019 1-year American 
Community Survey Data Profile, variable `DP03_0021P`. 
To determine this information, we can use *tidyverse* tools to sort our data by 
descending order of a summary variable representing total population and then 
retaining the 20 largest metropolitan areas by population using the `slice_max()` 
function.

```{r}
library(tidycensus)
library(tidyverse)

metros <- get_acs(
  geography = "cbsa",
  variables = "DP03_0021P",
  summary_var = "B01003_001",
  survey = "acs1",
  year = 2019
) %>% 
  slice_max(summary_est, n = 20)
```

The returned data frame has 7 columns, as is standard for `get_acs()` with a 
summary variable, but has 20 rows as specified by the `slice_max()` command. 
While the data can be filtered and sorted further to facilitate comparative 
analysis, it also can be represented succinctly with a visualization. 
The tidy format returned by `get_acs()` is well-suited for visualization with *ggplot2*.

In the basic example below, we can create a bar chart comparing public 
transportation as commute share for the most populous metropolitan areas in the 
United States with a minimum of code. The first argument to `ggplot()` in the 
example below is the name of our dataset; the second argument is an aesthetic 
mapping of columns to plot elements, specified inside the `aes()` function. 
This plot initialization is then linked with the `+` operator to the `geom_col()` 
function to create a bar chart.

```{r}
ggplot(metros, aes(x = NAME, y = estimate)) + geom_col()
```

While the above chart is a visualization of the `metros` dataset, 
it tells us little about the data given the lack of necessary formatting. 
The x-axis labels are so lengthy that they overlap and are impossible to read; 
the axis titles are not intuitive; and the data are not sorted, 
making it difficult to compare similar observations.

## 4.2.1 Improving plot legibility
Fortunately, the plot can be made more legible by cleaning up the metropolitan 
area name, re-ordering the data in descending order, then adding layers to the 
plot definition. Additionally, *ggplot2* visualization can be used in combination 
with *magrittr* piping and *tidyverse* functions, allowing analysts to string 
together data manipulation and visualization processes.

Our first step will be to format the `NAME` column in a more intuitive way. 
The `NAME` column by default provides a description of each geography as formatted 
by the US Census Bureau. However, a detailed description like 
`"Atlanta-Sandy Springs-Roswell, GA Metro Area"` is likely unnecessary for our chart, 
as the same metropolitan area can be represented on the chart by the name of its 
first principal city, which in this case would be `"Atlanta"`. To accomplish this, 
we can overwrite the `NAME` column by using the tidyverse function `str_remove()`, 
found in the *stringr* package. The example uses _regular expressions_ to first 
remove all the text after the first dash, then remove the text after the first
comma if no dash was originally present. These two subsequent calls to `mutate()`
will account for the various ways that metropolitan area names are specified.

On the chart, the legibility can be further improved by mapping the metro name 
to the y-axis and the ACS estimate to the x-axis, and plotting the points in 
descending order of their estimate values. The ordering of points in this way 
is accomplished with the `reorder()` function, used inside the call to `aes()`. 
As the result of the `mutate()` operations is piped to the `ggplot()` function in 
this example with the %>% operator, the dataset argument to `ggplot()` is inferred
by the function.

```{r}
metros %>% 
  mutate(NAME = str_remove(str_remove(NAME, "-.*$"), ",.*$")) %>% 
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) + geom_col()
```

The plot is much more legible after our modifications. 
Metropolitan areas can be directly compared with one another, and the metro area 
labels convey enough information about the different places without overwhelming 
the plot with long axis labels. However, the plot still lacks information to 
inform the viewer about the plot’s content. This can be accomplished by specifying 
_labels_ inside the `labs()` function. In the example below, we’ll specify a title and 
subtitle, and modify the X and Y axis labels from their defaults.

```{r}
metros %>% 
  mutate(NAME = str_remove(str_remove(NAME, "-.*$"), ",.*$")) %>% 
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) + geom_col() +
  theme_minimal() +
  labs(
    title = "Public transit commute share",
    subtitle = "2019 1-year ACS estimate",
    y = "",
    x = "ACS estimate",
    caption = "Source: ACS Data Profile variable DP03_0021P via the tidycensus R package"
  )
```

The inclusion of labels provides key information about the contents of the plot 
and also gives it a more polished look for presentation.

## 4.2.2 Custom styling of ggplot2 charts

While an analyst may be comfortable with the plot as-is, *ggplot2* allows for 
significant customization with respect to stylistic presentation. 
The example below makes a few such modifications. This includes styling the 
bars on the plot with a different color and internal transparency; 
changing the font; and customizing the axis tick labels.

```{r}
library(scales)

metros %>% 
  mutate(NAME = str_remove(str_remove(NAME, "-.*$"), ",.*$")) %>% 
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) + 
  geom_col(color = "navy", fill = "navy",
           alpha = 0.5, width = 0.85) +
  theme_minimal(base_size = 12, base_family = "Verdana") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Public transit commute share",
    subtitle = "2019 1-year ACS estimate",
    y = "",
    x = "ACS estimate",
    caption = "Source: ACS Data Profile variable DP03_0021P via the tidycensus R package"
  )
```

The code used to produced the styled graphic uses the following modifications:

* While aesthetic mappings relative to a column in the input dataset will be 
specified in a call to `aes()`, *ggplot2* geoms can be styled directly in their 
corresponding functions. Bars in *ggplot2* are characterized by both a __color__, 
which is the outline of the bar, and a __fill__. The above code sets both to `"navy"` 
then modifies the internal transparency of the bar with the `alpha` argument. 
Finally, `width = 0.85` slightly increases the spacing between bars.

* In the call to `theme_minimal()`, `base_size` and `base_family` parameters are available. 
`base_size` specifies the base font size to which plot text elements will be drawn; 
this defaults to 11. In many cases, you will want to increase `base_size` to improve plot legibility. 
`base_family` allows you to change the font family used on your plot. 
In this example, `base_family` is set to `"Verdana"`, but you can use any font 
families accessible to R from your operating system. 
To check this information, use the `system_fonts()` function in the *systemfonts* 
package (Pedersen, Ooms, and Govett 2021).

* The `scale_x_continuous()` function is used to customize the X-axis of the plot. 
The `labels` parameter can accept a range of values including a function (used here) 
or formula that operates over the tick labels. The *scales* 
([Wickham and Seidel, n.d.](https://walker-data.com/census-r/references.html#ref-scales)) 
package contains many useful formatting functions to 
neatly present tick labels, such as `label_percent()`, `label_dollar()`, and `label_date()`. 
The functions also accept arguments to modify the presentation.

## 4.2.3 Exporting data visualizations from R

Once an analyst has settled on a visualization design, 
they may want to export their image from R to display on a website, 
in a blog post, or in a report. Plots generated in RStudio can be exported with 
the *Export > Save as Image* command; however, analysts who want more programmatic 
control over their image exports can script this with *ggplot2*. 
The `ggsave()` function in *ggplot2* will save the last plot generated to an image 
file in the user’s current working directory by default. The specified file 
extension will control the output image format, e.g. `.png`.

```{r}
ggsave("metro_transit.png")
```

`ggsave()` includes a variety of options for more fine-grained control over the image output. 
Common options used by analysts will be `width` and `height` to control the image size; 
`dpi` to control the image resolution (in dots per inch); and `path` to specify the 
directory in which the image will be located. For example, the code below would 
write the most recent plot generated to an 8 inch by 5 inch image file in a 
custom location with a resolution of 300 dpi.

```{r}
ggsave(
  filename = "metro_transit.png",
  path = "../../imgs",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300
)
```

