# [4.3 Visualizing margins of error](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#visualizing-margins-of-error)

While *tidycensus* has tools available for working with margins of error in a data 
wrangling workflow, it is also often useful to visualize those margins of error 
to illustrate the degree of uncertainty around estimates, especially when making 
comparisons between those estimates.

In the example visualization in section 4.2 of public transportation mode share 
by metropolitan area for the largest metros in the United States, estimates are 
associated with margins of error; however, these margins of error are relatively 
small given the large population size of the geographic units represented in the 
plot. However, if studying demographic trends for geographies of smaller 
population size - like counties, Census tracts, or block groups - comparisons 
can be subject to a considerable degree of uncertainty.

## 4.3.1 Data setup

In the example below, we will compare the median household incomes of counties 
in the US state of Maine from the 2016-2020 ACS. Before doing so, it is helpful 
to understand some basic information about counties in Maine, such as the number 
of counties and their total population. We can retrieve this information with 
tidycensus and 2020 decennial Census data.

```{r}
library(tidycensus)
library(tidyverse)

maine <- get_decennial(
  state = "Maine",
  geography = "county",
  variables = c(totalpop = "P1_001N"),
  year = 2020
) %>% 
  arrange(desc(value))

maine
```

There are sixteen counties in Maine, ranging in population from a maximum of 
303,069 to a minimum of 16,800. In turn, estimates for the counties with smaller 
population sizes are likely to be subject to a larger margin of error than those 
with larger baseline populations.

Comparing median household incomes of these sixteen counties illustrates this point. 
Let’s first obtain this data with *tidycensus* then clean up the `NAME` column with 
`str_remove()` to remove redundant information.

```{r}
maine_income <- get_acs(
  state = "Maine",
  geography = "county",
  variables = c(hhincome = "B19013_001")
) %>% 
  mutate(NAME = str_remove(NAME, " County, Maine"))
```

Using some of the tips covered in the previous visualization section, 
we can produce a plot with appropriate styling and formatting to rank the counties.

```{r}
library(scales)

ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_point(size = 3, color = "darkgreen") +
  labs(
    title = "Median household income",
    subtitle = "Counties in Maine",
    x = "",
    y = "ACS estimate"
  ) +
  theme_minimal(base_size = 12.5) +
  scale_x_continuous(labels = label_dollar())
```

The above visualization suggests a ranking of counties from the wealthiest 
(Cumberland) to the poorest (Piscataquis). However, the data used to generate 
this chart is significantly different from the metropolitan area data used in 
the previous example. In our first example, ACS estimates covered the top 20 US 
metros by population - areas that all have populations exceeding 2.8 million. 
For these areas, margins of error are small enough that they do not meaningfully 
change the interpretation of the estimates given the large sample sizes used to 
generate them. However, as discussed in Section [3.5](https://walker-data.com/census-r/wrangling-census-data-with-tidyverse-tools.html#handling-margins-of-error-in-the-american-community-survey-with-tidycensus), smaller geographies may 
have much larger margins of error relative to their ACS estimates.

## 4.3.2 Using error bars for margins of error

Several county estimates on the chart are quite close to one another, 
which may mean that the ranking of counties is misleading given the margin of 
error around those estimates. We can explore this by looking directly at the 
data.

```{r}
maine_income %>% 
  arrange(desc(moe))
```

Specifically, margins of error around the estimated median household incomes 
vary from a low of \$1563 (Cumberland County) to a high of \$4616 (Sagadahoc County). 
In many cases, the margins of error around estimated county household income 
exceed the differences between counties of neighboring ranks, suggesting 
uncertainty in the ranks themselves.

In turn, a dot plot like the one above intended to visualize a ranking of county 
household incomes in Maine may be misleading. However, using visualization tools 
in *ggplot2*, we can visualize the uncertainty around each estimate, giving chart 
readers a sense of the uncertainty in the ranking. This is accomplished with the 
`geom_errorbar()` function, which will plot horizontal error bars around each dot 
that stretch to a given value around each estimate. In this instance, we will 
use the `moe` column to determine the lengths of the error bars.

```{r}
ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(size = 3, color = "darkgreen") +
  theme_minimal(base_size = 12.5) +
  labs(title = "Median household income",
       subtitle = "Counties in Maine",
       x = "2016-2020 ACS estimate",
       y = "") +
  scale_x_continuous(labels = label_dollar())
```

Adding the horizontal error bars around each point gives us critical information 
to help us understand how our ranking of Maine counties by median household income. 
For example, while the ACS estimate suggests that Piscataquis County has the 
lowest median household income in Maine, the large margin of error around the 
estimate for Piscataquis County suggests that either Aroostook or Washington 
Counties could conceivably have lower median household incomes. Additionally, 
while Hancock County has a higher estimated median household income than Lincoln, 
Waldo, and Knox Counties, the margin of error plot shows us that this ranking is 
subject to considerable uncertainty.
