# [Visualizing ACS estimates over time](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#visualizing-acs-estimates-over-time)

Section [3.4.2](https://walker-data.com/census-r/wrangling-census-data-with-tidyverse-tools.html#preparing-time-series-acs-estimates) covered how to obtain a time series of ACS estimates to explore 
temporal demographic shifts. While the output table usefully represented the 
time series of educational attainment in Colorado counties, data visualization 
is also commonly used to illustrate change over time. Arguably the most common 
chart type chosen for time-series visualization is the line chart, which *ggplot2* 
handles capably with the `geom_line()` function.

For an illustrative example, we’ll obtain 1-year ACS data from 2005 through 2019 
on median home value for Deschutes County, Oregon, home to the city of Bend and 
large numbers of in-migrants in recent years from the Bay Area in California. 
As in Chapter 3, `map_dfr()` is used to iterate over a named vector of years, 
creating a time-series dataset of median home value in Deschutes County since 2005, 
and we use the formula specification for anonymous functions so that `~ .x` 
translates to `function(x) x`.

```{r}
years <- 2005:2019
names(years) <- years

deschutes_value <- list_rbind(
  map(years, ~{
    get_acs(
      geography = "county",
      variables = "B25077_001",
      state = "OR",
      county = "Deschutes",
      year = .x,
      survey = "acs1"
    )
  }), names_to = "year"
  ) 

deschutes_value
```

This information can be visualized with familiar *ggplot2* syntax. `deschutes_value` 
is specified as the input dataset, with `year` mapped to the X-axis and `estimate` 
mapped to the y-axis. The argument `group = 1` is used to help *ggplot2* understand 
how to connect the yearly data points with lines given that only one county is 
being visualized. `geom_line()` then draws the lines, and we layer points on top 
of the lines as well to highlight the actual ACS estimates.

```{r}
ggplot(deschutes_value, aes(x = year, y = estimate, group = 1)) +
  geom_line() +
  geom_point()
```

The chart shows rising home values prior to the 2008 recession; 
a notable drop after the housing market crash; and rising values since 2011, 
reflecting increased demand from wealthy in-migrants from locations like the Bay Area. 
Given what we have learned in previous sections, there are also several opportunities 
for chart cleanup. This can include more intuitive tick and axis labels; 
a re-designed visual scheme; and a title and caption. We can also build the 
margin of error information into the line chart like we did in the previous section. 
We’ll use the *ggplot2* function `geom_ribbon()` to draw the margin of error interval 
around the line, helping represent uncertainty in the ACS estimates.

```{r}
ggplot(deschutes_value, aes(x = year, y = estimate, group = 1)) +
  geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe),
              fill = "navy",
              alpha = 0.4) +
  geom_line(color = "navy") +
  geom_point(color = "navy", size = 2) +
  theme_minimal(base_size = 12) +
  scale_y_continuous(labels = label_dollar(scale = .001, suffix = "k")) +
  labs(
    title = "Median home value in Deschutes County, OR",
    x = "Year",
    y = "ACS estimate",
    caption = "Shaded area represents margin of error around the ACS estimate"
  )
```



























