# [4.5 Exploring age and sex structure with population pyramids](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#visualizing-acs-estimates-over-time)

A common method for visualizing the demographic structure of a particular area 
is the __population pyramid__. Population pyramids are typically constructed by 
visualizing population size or proportion on the x-axis; age cohort on the y-axis; 
and sex is represented categorically with male and female bars mirrored around a 
central axis.

## 4.5.1 Preparing data from the Population Estimates API

We can illustrate this type of visualization using data from the 
Population Estimates API for the state of Utah. 
We first obtain data using the `get_estimates()` function in *tidycensus* for 
2019 population estimates from the Census Bureau’s Population Estimates API.

```{r}
utah <- get_estimates(
  geography = "state",
  state = "UT",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = T,
  year = 2019
)

utah
```

The function returns a long-form dataset in which each row represents population 
values broken down by age and sex for the state of Utah. However, there are some 
key issues with this dataset that must be addressed before constructing a 
population pyramid. First, several rows represent values that we don’t need for
our population pyramid visualization. For example, the first few rows in the
dataset represent population values for "Both sexes" or for "All ages". 
In turn, it will be necessary to isolate those rows that represent five-year age
bands by sex, and remove the rows that do not. This can be resolved with some
data wrangling using tidyverse tools.

In the dataset returned by `get_estimates()`, five-year age bands are identified 
in the `AGEGROUP` column beginning with the word `"Age"`. We can filter this dataset
for rows that match this pattern, and remove those rows that represent both sexes. 
This leaves us with rows that represent five-year age bands by sex. 
However, to achieve the desired visual effect, data for one sex must mirror 
another, split by a central vertical axis. To accomplish this, we can set the
values for all `Male` values to negative.

```{r}
utah_filtered <- filter(utah, str_detect(AGEGROUP, "^Age"),
                        SEX != "Both sexes") %>% 
  mutate(value = ifelse(SEX == "Male", -value, value))

utah_filtered
```

## 4.5.2 Designing and styling the population pyramid

The data are now ready for visualization. The core components of the pyramid 
visualization require mapping the population value and the age group to the chart
axes. Sex can be mapped to the `fill` aesthetic allowing for the plotting of these
categories by color

```{r}
ggplot(utah_filtered, aes(x = value, y = AGEGROUP, fill = SEX)) +
  geom_col()
```

The visualization represents a functional population pyramid that is nonetheless 
in need of some cleanup. In particular, the axis labels are not informative; 
the y-axis tick labels have redundant information (“Age” and “years”); 
and the x-axis tick labels are difficult to parse. Cleaning up the plot allows 
us to use some additional visualization options in *ggplot2*. In addition to 
specifying appropriate chart labels, we can format the axis tick labels by using
appropriate `scale_*` functions in *ggplot2* and setting the X-axis limits to show
both sides of 0 equally. In particular, this involves the use of custom absolute
values to represent population sizes, and the removal of redundant age group
information. We’ll also make use of an alternative *ggplot2* theme, 
`theme_minimal()`, which uses a white background with muted gridlines.

```{r}
utah_pyramid <- ggplot(
  utah_filtered,
  aes(
    x = value,
    y = AGEGROUP,
    fill = SEX
  )
) +
  geom_col(width = 0.95, alpha = 0.75) +
  theme_minimal(base_family = "Verdana",
                base_size = 12) +
  scale_x_continuous(
    labels = ~number_format(scale = .001, suffix = "k")(abs(.x)),
    limits = 140000 * c(-1,1)
  ) +
  scale_y_discrete(labels = ~str_remove_all(.x, "Age\\s|\\syears")) +
  scale_fill_manual(values = c("darkred", "navy")) +
  labs(
    x = "",
    y = "2019 Census Bureau population estimate",
    title = "Population structure in Utah",
    fill = "",
    caption = "Data source: US Census Bureau population estimates & tidycensus R package"
    )

utah_pyramid
```

















