# [4.6 Visualizing group-wise comparison](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#visualizing-group-wise-comparisons)

One of the most powerful features of **ggplot2** is its ability to generate 
_faceted_ plots, which are also commonly referred to as _small multiples_. 
Faceted plots allow for the sub-division of a dataset into groups, 
which are then plotted side-by-side to facilitate comparisons between those groups. 
This is particularly useful when examining how distributions of values vary 
across different geographies. An example shown below involves a comparison of 
median home values by Census tract for three counties in the Portland, Oregon area: 
Multnomah, which contains the city of Portland, 
and the suburban counties of Clackamas and Washington.

```{r}
housing_val <- get_acs(
  geography = "tract",
  variables = "B25077_001",
  state = "OR",
  county = c(
    "Multnomah",
    "Clackamas",
    "Washington",
    "Yamhill",
    "Marion",
    "Columbia"
  ),
  year = 2020
)

housing_val
```

As with other datasets obtained with **tidycensus**, the `NAME` column contains 
descriptive information that can be parsed to make comparisons. 
In this case, Census tract ID, county, and state are separated with commas; 
in turn the tidyverse `separate()` function can split this column into three 
columns accordingly.

```{r}
housing_val2 <- separate(
  housing_val,
  NAME,
  into = c("tract", "county", "state"),
  sep = ", "
)
housing_val2
```

As explored in previous chapters, 
a major strength of the tidyverse is its ability to perform group-wise data analysis. 
The dimensions of median home values by Census tract in each of the three 
counties can be explored in this way. For example, a call to `group_by()` followed 
by `summarize()` facilitates the calculation of county minimums, means, medians, and maximums.

```{r}
housing_val2 %>% 
  group_by(county) %>% 
  summarize(min = min(estimate, na.rm = T),
            mean = mean(estimate, na.rm = T),
            median = median(estimate, na.rm = T),
            max = max(estimate, na.rm = T))
```

While these basic summary statistics offer some insights into comparisons between 
the three counties, they are limited in their ability to help us understand the 
dynamics of the overall distribution of values. This task can in turn be 
augmented through visualization, which allows for quick visual comparison of 
these distributions. Group-wise visualization in **ggplot2** can be accomplished 
with the `facet_wrap()` function added onto any existing **ggplot2** code that has 
salient groups to visualize. For example, a kernel density plot can show the 
overall shape of the distribution of median home values in our dataset:


```{r}
ggplot(housing_val2, aes(x = estimate)) + geom_density()
```
Mapping the `county` column onto the `fill` aesthetic will then draw superimposed 
density plots by county on the chart:

```{r}
ggplot(housing_val2, aes(x = estimate, fill = county)) + geom_density(alpha = 0.3)
```

Alternatively, adding the `facet_wrap()` function, and specifying `county` as 
the column used to group the data, splits this visualization into side-by-side
graphics based on the counties to which each Census tract belongs.

```{r}
ggplot(housing_val2, aes(x = estimate)) + 
  geom_density(fill = "darkgreen", color = "darkgreen", alpha = 0.5) +
  facet_wrap(~county) +
  scale_x_continuous(labels = dollar_format(scale = 0.000001, suffix = "m")) +
  theme_minimal(base_size = 14) +
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(angle = 45)) +
  labs(x = "ACS estimate",
       y = "",
       title = "Median home values by Census tract, 2016-2020 ACS") 
```

The side-by-side comparative graphics show how the value distributions vary 
between the three counties. Home values in all three counties are common around
$250,000, but Multnomah County has some Census tracts that represent the highest
values in the dataset.
