# [4.7 Advanced visualization with ggplot2 extension](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#advanced-visualization-with-ggplot2-extensions)

While the core functionality of **ggplot2** is very powerful, a notable advantage
of using **ggplot2** for visualization is the contributions made by its user 
community in the form of _extensions_. **ggplot2** extensions are packages developed 
by practitioners outside the core **ggplot2** development team that add functionality 
to the package. I encourage you to [review the gallery of ggplot2 extensions from 
the extensions website](https://exts.ggplot2.tidyverse.org/); 
I highlight some notable examples below.

## ggridges

The **ggridges** package adapts the concept of the faceted density plot to 
generate _ridgeline plots_, in which the densities overlap one another. 
The example below creates a ridgeline plot using the Portland-area home value data; 
`geom_density_ridges()` generates the ridgelines, and `theme_ridges()` styles the 
plot in an appropriate manner.

```{r}
library(ggridges)

ggplot(housing_val2, aes(x = estimate, y = county)) +
  geom_density_ridges() +
  theme_ridges() +
  labs(x = "Median home value: 2016-2020 ACS estimate",
       y = "") +
  scale_x_continuous(labels = label_dollar(scale = .000001, suffix = "m"),
                     breaks = c(0, 5e5, 1e6)) +
  theme(axis.text.x = element_text(angle = 45))
```

The overlapping density “ridges” offer both a pleasing aesthetic but also a 
practical way to compare the different data distributions. As **ggridges** extends 
**ggplot2**, analysts can style the different chart components to their liking using 
the methods introduced earlier in this chapter.

## 4.7.2 ggbeeswarm

The **ggbeeswarm** package extends

ggplot2 by allowing users to generate _beeswarm_ plots, in which clouds of points 
are jittered to show the overall density of a distribution of data values. 
Beeswarm plots can be compelling ways to visualize multiple data variables on a chart, 
such as the distributions of median household income by the racial and ethnic 
composition of neighborhoods. This is the motivating example for the chart below, 
which looks at household income by race/ethnicity in New York City. 
The data wrangling in the first part of the code chunk takes advantage of some of 
the skills covered in Chapters 2 and 3, allowing for visualization of household 
income distributions by the largest group in each Census tract.

```{r}
library(ggbeeswarm)

nyc_race_income <- get_acs(
  geography = "tract",
  state = "NY",
  county = c("New York", "Bronx", "Queens", "Richmond", "Kings"),
  variables = c(White = "B03002_003",
                Black = "B03002_004",
                Asian = "B03002_006",
                Hispanic = "B03002_012"),
  summary_var = "B19013_001",
  year = 2022
) %>% 
  group_by(GEOID) %>% 
  filter(estimate == max(estimate, na.rm = T)) %>% 
  ungroup() %>% 
  filter(estimate != 0)

ggplot(nyc_race_income, aes(x = variable, y = summary_est, color = summary_est)) + 
  geom_quasirandom(alpha = 0.5) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  scale_color_viridis_c(guide = "none") +
  scale_y_continuous(labels = label_dollar()) +
  labs(x = "Largest group in Census tract",
       y = "Median household income",
       title = "Household income distribution by largest racial/ethnic group",
       subtitle = "Census tracts, New York City",
       caption = "Data source: 2018-2022 ACS")
```

The plot shows that the wealthiest neighborhoods in New York City - 
those with median household incomes exceeding $150,000 - are nearly all 
plurality or majority non-Hispanic white. However, the chart also illustrates 
that there are a range of values among neighborhoods with pluralities of the 
different racial and ethnic groups, suggesting a nuanced portrait of the 
intersections between race and income inequality in the city.

## 4.7.3 Geofaced plots

The next four chapters of the book, Chapters 5 through 8, 
are all about spatial data, mapping, and spatial analysis. 
Geographic location can be incorporated into Census data visualizations without
using geographic information explicitly by way of the **geofacet** package. 
Geofaceted plots are enhanced versions of faceted visualizations that arrange
subplots in relationship to their relative geographic location.
The geofacet package has [over 100 available grids to choose from](https://hafen.github.io/geofacet/articles/geofacet.html#list-available-grids) allowing for
faceted plots for US states, counties, and regions around the world.
The key is to use a column that can map correctly to information in the
geofaceted grid that you are using.

In the example below, we replicate the population pyramid code above to generate
population pyramids for each state in the US. However, we also modify the output
data so that the population information reflects the proportion of the overall
population for each state so that all states are on consistent scales.
The population pyramid code is similar, though the axis information is
unnecessary as it will be stripped from the final plot.

```{r}
library(geofacet)

us_pyramid_data <- get_estimates(
  geography = "state",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = T,
  year = 2022
) %>% 
  filter(str_detect(AGEGROUP, "^Age"),
         SEX != "Both sexes") %>% 
  group_by(NAME) %>% 
  mutate(prop = value / sum(value, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(prop = ifelse(SEX == "Male", -prop, prop))

ggplot(us_pyramid_data, aes(x = prop, y = AGEGROUP, fill = SEX)) +
  geom_col(width = 1) +
  theme_minimal() +
  scale_fill_manual(values = c("darkred", "navy")) +
  facet_geo(~NAME, grid = "us_state_with_DC_PR_grid2",
            label = "code") +
  theme(axis.text = element_blank(),
        strip.text.x = element_text(size = 8)) +
  labs(x = "",
       y = "",
       title = "Population structure by age and sex",
       fill = "",
       caption = "Data source: US Census Bureau population estimates & tidycensus R package")

```

The result is a compelling visualization that expresses general geographic 
relationships and patterns in the data while showing comparative population 
pyramids for all states. The unique nature of Washington, DC stands out with its
large population in their 20s and 30s; we can also compare Utah, the youngest
state in the country, with older states in the Northeast.

If this specific grid arrangement is not to your liking 
(e.g. Minnesotans may take issue with Wisconsin to the north!), try out some of
the other built-in grid options in the package. Also, you can use the
`grid_design()` function in the geofacet package to pull up an interactive app
in which you can design your own grid for use in geofaceted plots!

## 4.7.4 Interactive visualization with plotly

The **htmlwidgets** package provides a bridge between JavaScript libraries for 
interactive web-based data visualization and the R language. 
Since 2015, the R community has released hundreds of packages that depend on
htmlwidgets and bring interactive data visualization to R. 
One of the most popular packages for interactive visualization is the **plotly** package,
an interface to the plotly visualization library.

**plotly** is a well-developed library that allows for many different types of
custom visualizations. However, one of the most useful functions in the **plotly**
package is arguably the simplest. The `ggplotly()` function can convert an
existing **ggplot2** graphic into an interactive web visualization in a single line
of code! Let’s try it out here with the `utah_pyramid` population pyramid:

```{r}
library(plotly)

utah <- get_estimates(
  geography = "state",
  state = "UT",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = T,
  year = 2019
)

utah_filtered <- filter(utah, str_detect(AGEGROUP, "^Age"),
                        SEX != "Both sexes") %>% 
  mutate(value = ifelse(SEX == "Male", -value, value))

utah_pyramid <- ggplot(
  utah_filtered,
  aes(
    x = value,
    y = AGEGROUP,
    fill = SEX
  )
) +
  geom_col(width = 0.95, alpha = 0.75) +
  theme_minimal(base_family = "Verdana",
                base_size = 12) +
  scale_x_continuous(
    labels = ~number_format(scale = .001, suffix = "k")(abs(.x)),
    limits = 140000 * c(-1,1)
  ) +
  scale_y_discrete(labels = ~str_remove_all(.x, "Age\\s|\\syears")) +
  scale_fill_manual(values = c("darkred", "navy")) +
  labs(
    x = "",
    y = "2019 Census Bureau population estimate",
    title = "Population structure in Utah",
    fill = "",
    caption = "Data source: US Census Bureau population estimates & tidycensus R package"
    )

ggplotly(utah_pyramid)
```

Try hovering your cursor over the different bars; this reveals a tooltip that 
shares information about the data. The legend is interactive, as data series can
be clicked on and off; viewers can also pan and zoom on the plot using the toolbar
that appears in the upper right corner of the visualization. Interactive graphics
like this are an excellent way to facilitate additional data exploration, and
can be polished further for presentation on the web.

# [4.8 Learning more about visualization](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#learning-more-about-visualization)

This chapter has introduced a series of visualization techniques implemented in
ggplot2 that are appropriate for US Census Bureau data. Readers may want to learn
more about effective principles for visualization design and communication and
apply that to the techniques covered here. While Chapter [6](https://walker-data.com/census-r/mapping-census-data-with-r.html#mapping-census-data-with-r) covers some of these
principles in brief, literature that focuses specifically on these topics will
be more comprehensive. Munzner ([2014](https://walker-data.com/census-r/references.html#ref-munzner2014)) is an in-depth overview of visualization
techniques and design principles, and offers a [corresponding website with lecture slides](https://www.cs.ubc.ca/~tmm/vadbook/). Evergreen ([2020](https://walker-data.com/census-r/references.html#ref-evergreen2020)) and Knaflic ([2015](https://walker-data.com/census-r/references.html#ref-knaflic2015)) provide guidelines for effective
communication with visualization and both offer excellent design tips for
business and general audiences. R users may be interested in Wilke ([2019](https://walker-data.com/census-r/references.html#ref-wilke2019)) and
Healy ([2019](https://walker-data.com/census-r/references.html#ref-healy2019)), which both offer a comprehensive overview of data visualization best
practices along with corresponding R code to reproduce the figures in their books.
