# [Programming with dplyr](https://dplyr.tidyverse.org/articles/programming.html)

Most dplyr verbs use tidy evaluation in some way. Tidy evaluation is a special type of non-standard evaluation used throughout the tidyverse. There are two basic forms found in dplyr:

`arrange()`, `count()`, `filter()`, `group_by()`, `mutate()`, and `summarise()` use _data masking_ so that you can use data variables as if they were variables in the environment (i.e. you write `my_variable` not `df$my_variable`).

`across()`, `relocate()`, `rename()`, `select()`, and `pull()` use _tidy selection_ so you can easily choose variables based on their position, name, or type (e.g. `starts_with("x")` or `is.numeric`).

If you’d like to learn more about the underlying theory, or precisely how it’s different from non-standard evaluation, we recommend that you read the Metaprogramming chapters in [Advanced R](https://adv-r.hadley.nz/metaprogramming.html).

## Data masking

In most (but not all1) base R functions you need to refer to variables with `$`, leading to code that repeats the name of the data frame many times:

```{r}starwars[starwars$homeworld == "Naboo" & starwars$species == "Human", ,]```

The dplyr equivalent of this code is more concise because data masking allows you to need to type the `starwars` dataframe once:

```{r}starwars %>% filter(homeworld == "Naboo", species == "Human")```

Note that the dplyr method automatically removes `NA` values while the base R method does not.

### Data- and env-variables

The key idea behind data masking is that it blurs the line between the two different meanings of the word “variable”:

env-variables are “programming” variables that live in an environment. They are usually created with `<-`.

data-variables are “statistical” variables that live in a data frame. They usually come from data files (e.g. `.csv`, `.xls`), or are created manipulating existing variables.

Take this piece of code:

```{r}
df <- data.frame(x = runif(3), y = runif(3))
df$x
```
It creates a env-variable, `df`, that contains two data-variables, `x` and `y`. 
Then it extracts the data-variable `x` out of the env-variable df using `$`.

This blurring of the meaning of “variable” is a really nice feature for interactive data analysis because it allows you to refer to data-vars as is, without any prefix. 

Unfortunately, this benefit does not come for free. When you start to program with these tools, you’re going to have to grapple with the distinction.

### Data masking indirection

The main challenge of programming with functions that use data masking arises when you introduce some indirection, i.e. when you want to get the data-variable from an env-variable instead of directly typing the data-variable’s name. There are two main cases:

* When you have the data-variable in a function argument (i.e. an env-variable that holds a promise), you need to **embrace** the argument by surrounding it in doubled braces, like `filter(df, {{ var }})`. The following function uses embracing to create a wrapper around `summarise()` that computes the minimum and maximum values of a variable, as well as the number of observations that were summarised:

```{r}
var_summary <- function(data, var) {
  data %>%
    summarise(n = n(), min = min({{ var }}), max = max({{ var }}))
}
mtcars %>% 
  group_by(cyl) %>% 
  var_summary(mpg)
```

* When you have an env-variable that is a character vector, you need to index into the `.data` pronoun with `[[`, like `summarise(df, mean = mean(.data[[var]]))`.

The following example uses `.data` to count the number of unique values in each variable of `mtcars`:

```{r}
for (var in names(mtcars)) {
  mtcars %>% count(.data[[var]]) %>% print()
}
```
Note that `.data` is not a data frame; it’s a special construct, a pronoun, that allows you to access the current variables either directly, with `.data$x` or indirectly with `.data[[var]]`. Don’t expect other functions to work with it.

### Name injection

Many data masking functions also use dynamic dots, which gives you another useful feature: generating names programmatically by using `:=` instead of `=`. There are two basics forms, as illustrated below with `tibble()`:

* If you have the name in an env-variable, you can use _glue_ syntax to interpolate in:
```{r}
name <- "susan"
tibble("{name}" := 2)
```

* If the name should be derived from a data-variable in an arguement, you can use _embracing_ syntax:
```{r}
my_df <- function(x) {
  tibble("{{x}}_2" := x * 2)
}
my_var <- 10
my_df(my_var)
```
Learn more in [?rlang::`dyn-dots`](https://rlang.r-lib.org/reference/dyn-dots.html).

## Tidy selection
Tidy selection is a complementary tool that makes it easy to work with the columns of a dataset.

### The tidyselect DSL

Underneath all functions that use tidy selection is the tidyselect package. It provides a miniature domain specific language that makes it easy to select columns by name, position, or type. For example:

* `select(df, 1)` selects the first column; `select(df, last_col())` selects the last column.
* `select(df, c(a, b, c))` selects columns `a`, `b`, and `c`.
* `select(df, starts_with("a"))` selects all columns whose name starts with "a"; `select(df, ends_with("z"))` selects all columns whose name ends with "z".
* `select(df, where(is.numeric))` selects all numeric columns.

You can see more details in [?dplyr_tidy_select](https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html)

### Tidy selection indirection

As with data masking, tidy selection makes a common task easier at the cost of making a less common task harder. When you want to use tidy select indirectly with the column specification stored in an intermediate variable, you’ll need to learn some new tools. Again, there are two forms of indirection:

* When you have the data-variable in an env-variable that is a function argument, you use the same technique as data masking: you **embrace** the argument by surrounding it in doubled braces.

The following function summarises a data frame by computing the mean of all variables selected by the user:

```{r}
summarise_mean <- function(data, vars) {
  data %>% summarise(n = n(), across({{ vars }}, mean))
}
mtcars %>% 
  group_by(cyl) %>% 
  summarise_mean(where(is.numeric))
```

* When you have an env-variable that is a character vector, you need to use `all_of()` or `any_of()` depending on whether you want the function to error if a variable is not found.

The following code uses `all_of()` to select all of the variables found in a character vector; then `!` plus `all_of()` to select all of the variables not found in a character vector:

```{r}
vars <- c("mpg", "vs")
mtcars %>% select(all_of(vars))
mtcars %>% select(!all_of(vars))
```

## How-tos


