# [3.2 Exploring Census data with tidyverse tools](https://walker-data.com/census-r/wrangling-census-data-with-tidyverse-tools.html#exploring-census-data-with-tidyverse-tools)

To get started, the tidycensus and tidyverse packages are loaded. “tidyverse” is not specifically a package itself, but rather loads several core packages within the tidyverse. The package load message gives you more information

```{r}
library(tidycensus)
library(tidyverse)
```

## [3.2.1 Sorting and filtering data](https://walker-data.com/census-r/wrangling-census-data-with-tidyverse-tools.html#sorting-and-filtering-data)

Let’s request data on median age from the 2016-2020 ACS with `get_acs()` for all counties in the United States. This requires specifying `geography = "county"` and leaving state set to `NULL`, the default.

```{r}
median_age <- get_acs(
    geography = "county",
    variables = "B01002_001",
    year = 2020
)
```
A first exploratory data analysis question might involve understanding which counties are the _youngest_ and _oldest_ in the United States as measured by median age. This task can be accomplished with the `arrange()` function found in the **dplyr** package. `arrange()` sorts a dataset by values in one or more columns and returns the sorted result. To view the dataset in ascending order of a given column, supply the data object and a column name to the `arrange()` function.

```{r}
arrange(median_age, estimate)
```

To retrieve the oldest counties in the United States by median age, an analyst can use the `desc()` function available in **dplyr** to sort the estimate column in descending order.

```{r}
arrange(median_age, desc(estimate))
```

The `filter()` function in **dplyr** queries a dataset for rows where a given condition evaluates to `TRUE`, and retains those rows only. For analysts who are familiar with databases and SQL, this is equivalent to a `WHERE` clause. This helps analysts subset their data for specific areas by their characteristics, and answer questions like “how many counties in the US have a median age of 50 or older?”


```{r}
filter(median_age, estimate >= 50)
```

Functions like `arrange()` and `filter()` operate on row values and organize data by row. Other tidyverse functions, like **tidyr**’s `separate()`, operate on columns. The `NAME` column, returned by default by most **tidycensus** functions, contains a basic description of the location that can be more intuitive than the `GEOID`. For the 2016-2020 ACS, `NAME` is formatted as “X County, Y”, where X is the county name and Y is the state name. `separate()` can split this column into two columns where one retains the county name and the other retains the state; this can be useful for analysts who need to complete a comparative analysis by state.

```{r}
separate(
    median_age,
    NAME,
    into = c("county", "state"),
    sep = ", "
)
```

## [3.2.2 Using summary variables and calculating new columns](https://walker-data.com/census-r/wrangling-census-data-with-tidyverse-tools.html#using-summary-variables-and-calculating-new-columns)

Data in Census and ACS tables, as in the example above, are frequently comprised of variables that individually constitute sub-categories such as the numbers of households in different household income bands. One limitation of the approach above, however, is that the data and the resulting analysis return estimated counts, which are difficult to compare across geographies. For example, Maricopa County in Arizona is the state’s most populous county with 4.3 million residents; the second-largest county, Pima, only has just over 1 million residents and six of the state’s 15 counties have fewer than 100,000 residents. In turn, comparing Maricopa’s estimates with those of smaller counties in the state would often be inappropriate.

A solution to this issue might involve **normalizing** the estimated count data by dividing it by the overall population from which the sub-group is derived. Appropriate denominators for ACS tables are frequently found in the tables themselves as variables. In ACS table B19001, which covers the number of households by income bands, the variable `B19001_001` represents the total number of households in a given enumeration unit, which we removed from our analysis earlier. Given that this variable is an appropriate denominator for the other variables in the table, it merits its own column to facilitate the calculation of proportions or percentages.

In **tidycensus**, this can be accomplished by supplying a variable ID to the `summary_var` parameter in both the `get_acs()` and `get_decennial()` functions. When using `get_decennial()`, doing so will create two new columns for the decennial Census datasets, `summary_var` and `summary_value`, representing the summary variable ID and the summary variable’s value. When using `get_acs()`, using `summary_var` creates three new columns for the ACS datasets, `summary_var`, `summary_est`, and `summary_moe`, which include the ACS estimate and margin of error for the summary variable.

The following example uses the `summary_var` parameter to compare the population of counties in Arizona by race & Hispanic origin with their baseline populations, using data from the 2016-2020 ACS.

```{r}
race_vars <- c(
  White = "B03002_003",
  Black = "B03002_004",
  Native = "B03002_005",
  Asian = "B03002_006",
  HIPI = "B03002_007",
  Hispanic = "B03002_012"
)

az_race <- get_acs(
  geography = "county",
  state = "AZ",
  variables = race_vars,
  summary_var = "B03002_001",
  year = 2020
) 
```

By using dplyr’s `mutate()` function, we calculate a new column, `percent`, representing the percentage of each Census tract’s population that corresponds to each racial/ethnic group in 2016-2020. The `select()` function, also in dplyr, retains only those columns that we need to view.

```{r}
az_race_percent <- az_race %>%
    mutate(percent = 100 * (estimate / summary_est)) %>%
    select(NAME, variable, percent)
```
