[8.1 Indices of Segregation and Diversity](https://walker-data.com/census-r/modeling-us-census-data.html#indices-of-segregation-and-diversity)

Census data can also be used to derive models for explaining patterns that occur across regions or within cities. These models draw from concepts introduced in prior chapters, but can also be used as part of explanatory frameworks or within broader analytic pipelines for statistical inference or machine learning.

A large body of research in the social sciences is concerned with neighborhood segregation and diversity. Segregation as addressed here generally refers to the measurement of the extent to which two or more groups live apart from each other; diversity as a companion metric measures neighborhood heterogeneity among groups.

Segregation and diversity indices are implemented in a variety of different R packages; the package recommended by this book is the **segregation** package, which includes R functions for a [variety of regional and local indices](https://journals.sagepub.com/doi/full/10.1177/0049124121986204).

## 8.1.1 Data setup withi spatial analysis

Much of the segregation and diversity literature focuses on race and ethnicity, which will be explored in the example below. The data setup code uses spatial methods covered in the previous three chapters to acquire Census tract-level data on population estimates for non-Hispanic white, non-Hispanic black, non-Hispanic Asian, and Hispanic populations in California, then filters those Census tracts those that intersect the largest urbanized areas by population in the state using an inner spatial join. In turn, it is an illustrative example of how spatial analysis tools can be important parts of data setup workflows for analysis.

```{r}
library(tidycensus)
library(tidyverse)
library(segregation)
library(tigris)
library(sf)

# Get California tract data by race/ethnicity
ca_acs_data <- get_acs(
  geography = "tract",
  variables = c(
    white = "B03002_003",
    black = "B03002_004",
    asian = "B03002_006",
    hispanic = "B03002_012"
  ),
  state = "CA",
  geometry = T,
  year = 2019
)

# Use tidycensus to get urbanized areas by population with geometry,
# then filter for those that have populations of 750k or more
us_urban_areas <- get_acs(
  geography = "urban area",
  variables = "B01001_001",
  geometry = T,
  year = 2019,
  survey = "acs1"
) %>% 
  filter(estimate >= 750000) %>% 
  transmute(urban_name = str_remove(NAME, fixed(", CA Urbanized Area (2010)")))

# Compute an inner spatial join between California tracts and the urbanized
# areas, returning tracts in the largest California urban areas with the 
# urban_name column appended

ca_urban_data <- ca_acs_data %>% 
  st_join(us_urban_areas, left = FALSE) %>% 
  select(-NAME) %>% 
  st_drop_geometry()
```

To summarize, the spatial analysis workflow detailed above uses the following steps:

1. Data on race & ethnicity from the 2015-2019 5-year ACS for the four largest demographic groups in California is acquired with **tidycensus**’s `get_acs()` at the Census tract level with feature geometry included. Depending on the goals of the study, other racial/ethnic groups (e.g. native American, native Hawaiian/Pacific Islander) should be added or removed as needed.

2. As urban areas as defined by the Census Bureau often cross state boundaries, urban areas must be obtained for the entire US with `get_acs()`. Once obtained, urban areas are filtered to only those areas with populations of 750,000 or greater, and then `transmute()` is used to retain only a new column representing the area name (along with the simple feature geometry column).

3. A spatial join between the Census tract data and the urban area data is computed with `st_join()`. The argument `left = FALSE` computes an _inner spatial join_, which retains only those Census tracts that intersect the urban area boundaries, and appends the corresponding `urban_name` column to each Census tract.

The data structure appears as follows:

```ca_urban_data```

The data are in long (tidy) form, the default used by **tidycensus**; this data structure is ideal for computing indices in the **segregation** package.

## 8.1.2 The dissimilarity index

The dissimilarity index is widely used to assess neighborhood segregation between two groups within a region. It is computed as follows:

$D=\frac{1}{2} \displaystyle\sum_{i=1}^N\left|\frac{a_i}{A}-\frac{b_i}{B}\right|$

where $a_i$ represents the population of group $A$ in a given areal unit $i$; 
$A$ is the total population of that group in the study region (e.g. a metropolitan area);
and $b_i$ and $B$ are the equivalent metrics for the second group. The index ranges from a low of 0 to a high of 1, where 0 represents perfect integration between the two groups and 1 represents complete segregation. This index is implemented in the **segregation** package with the `dissimilarity()` function.

The example below computes the dissimilarity index between non-Hispanic white and Hispanic populations for the San Francisco/Oakland urbanized area. The data are filtered for only those rows that represent the target populations in the San Francisco/Oakland area, which is then piped to the `dissimilarity()` function. The function requires identification of a group column, for which we’ll use `variable`; a unit column representing the neighborhood unit, for which we’ll use `GEOID` to represent the Census tract; and a weight column that tells the function how many people are in each group.

```{r}
ca_urban_data %>% 
  filter(variable %in% c("white", "hispanic"),
         urban_name == "San Francisco--Oakland") %>% 
  dissimilarity(
    group = "variable",
    unit = "GEOID",
    weight = "estimate"
  )
```

The D index of segregation between non-Hispanic white and Hispanic populations in the San Francisco-Oakland area is 0.51. This statistic, however, is more meaningful in comparison with other cities. To compute dissimilarity for each urban area, we can creatively apply tidyverse techniques covered in earlier chapters and introduce a new function, `group_modify()`, for group-wise calculation. This example follows the recommended workflow [in the segregation package documentation](https://elbersb.github.io/segregation/articles/faq.html#how-can-i-compute-indices-for-different-areas-at-once-). The code below filters the data for non-Hispanic white and Hispanic populations by Census tract, then groups the dataset by values in the `urban_name` column. The `group_modify()` function from **dplyr** then allows for the calculation of dissimilarity indices _by group_, which in this example is Census tracts within each respective urban area. It returns a combined dataset that is sorted in descending order with `arrange()` to make comparisons.

```{r}
ca_urban_data %>% 
  filter(variable %in% c("white", "hispanic")) %>% 
  group_by(urban_name) %>% 
  group_modify(~dissimilarity(
    .x,
    group = "variable",
    unit = "GEOID",
    weight = "estimate"
  )) %>% 
  arrange(desc(est))
```
The Los Angeles area is the most segregated of the large urbanized areas in California with respect to non-Hispanic white and Hispanic populations at the Census tract level, followed by San Francisco/Oakland. Riverside/San Bernardino and Sacramento are the least segregated of the large urban areas in the state.

## 8.1.3 Multi-group segregation indices
One disadvantage of the dissimilarity index is that it only measures segregation between two groups. For a state as diverse as California, we may be interested in measuring segregation and diversity between multiple groups at a time. The **segregation** package implements two such indices: the Mutual Information Index $M$, and Theil’s Entropy Index $H$. $M$ is computed as follows for a dataset $T$:

$M(\mathbf{T})=\displaystyle\sum_{u=1}^U \sum_{g=1}^G p_{u g} \log \frac{p_{u g}}{p_u p_g}$

where $U$ is the total number of units $u$, $G$ is the total number of groups $g$, and $p_{ug}$ is the joint probability of being in unit $u$ and group $g$, with $p_u$ and $p_g$ referring to unit and group probabilities. Theil’s $H$ for the same dataset $T$ can then be written as:

$H(T) = \frac{M(T)}{E(T)}$

Where $E(T)$ is the entropy of $T$, normalizing $H$ to range between values of 0 and 1.

Computing these indices is straightforward with the **segregation** package. The `mutual_total()` function computes both indices; when different regions are to be considered (like multiple urban areas, as in this example) the `mutual_within()` function will compute $M$ and $H$ by urban area with the `within` argument appropriately specified. We’ll be using the full `ca_urban_data` dataset, which includes population estimates for non-Hispanic white, non-Hispanic Black, non-Hispanic Asian, and Hispanic populations.

```{r}
mutual_within(
  data = ca_urban_data,
  group = "variable",
  unit = "GEOID",
  weight = "estimate",
  within = "urban_name",
  wide = T
)
```
When multi-group segregation is considered using these indices, Los Angeles remains the most segregated urban area, whereas Riverside/San Bernardino is the least segregated.

The **segregation** package also offers a function for local segregation analysis, `mutual_local()`, which decomposes $M$
into unit-level segregation scores, represented by `ls`. In the example below, we will use `mutual_local()` to examine patterns of segregation across the most segregated urban area, Los Angeles.

```{r}
la_local_seg <- ca_urban_data %>% 
  filter(urban_name == "Los Angeles--Long Beach--Anaheim") %>% 
  mutual_local(
    group = "variable",
    unit = "GEOID",
    weight = "estimate",
    wide = T
  )
la_local_seg
```
The results can be mapped by joining the data to a dataset of Census tracts from **tigris**; the `inner_join()` function is used to retain tracts for the Los Angeles area only.

```{r}
la_tracts_seg <- tracts("CA", cb = T, year = 2019) %>% 
  inner_join(la_local_seg, by = "GEOID")

la_tracts_seg %>% 
  ggplot(aes(fill = ls)) +
  geom_sf(color = NA) +
  coord_sf(crs = 26946) +
  scale_fill_viridis_c(option = "inferno") +
  theme_void() +
  labs(fill = "Local\nsegregation index")
```
## 8.1.4 Visualizing the diversity gradient
The diversity gradient is a concept that uses scatterplot smoothing to visualize how neighborhood diversity varies by distance or travel-time from the core of an urban region. Historically, literature on suburbanization in the social sciences assumes a more heterogeneous urban core relative to segregated and homogeneous suburban neighborhoods. The diversity gradient is a visual heuristic used to evaluate the validity of this demographic model.

The entropy index for a given geographic unit is calculated as follows:
$E = \displaystyle\sum_{r=1}^nQ_rln\frac{1}{Q_r}$

$Q_r$ in this calculation represents group $r$'s proportion of the population in the geographic unit.

This statistic is implemented in the `entropy()` function in the **segregation** package. As the `entropy()` function calculates this statistic for a specific unit at a time, we will group the data by tract, and then use `group_modify()` to calculate the entropy for each tract separately. The argument `base = 4` is set by convention to the number of groups in the calculation; this sets the maximum value of the statistic to 1, which represents perfect evenness between the four groups in the area. Once computed, the indices are joined to a dataset of Census tracts from California; `inner_join()` is used to retain only those tracts in the Los Angeles urbanized area.

```{r}
la_entropy <- ca_urban_data %>% 
  filter(urban_name == "Los Angeles--Long Beach--Anaheim") %>% 
  group_by(GEOID) %>% 
  group_modify(~tibble(
    entropy = entropy(
      data = .x,
      group = "variable",
      weight = "estimate",
      base = 4
    )
  ))

la_entropy_geo <- tracts("CA", cb = T, year = 2019) %>% 
  inner_join(la_entropy, by = "GEOID")
```

Visualization of the diversity gradient then requires a relative measurement of how far each Census tract is from the urban core. The travel-time methods available in the mapboxapi package introduced in Chapter 7 are again used here to calculate driving distance to Los Angeles City Hall for all Census tracts in the Los Angeles urbanized area.

```{r}
library(mapboxapi)

la_city_hall <- mb_geocode("City Hall, Los Angeles CA")

mins_to_dwntwn <- mb_matrix(la_entropy_geo, la_city_hall)
```
Once computed, the travel times are stored in a vector `minutes_to_downtown`, then assigned to a new column `minutes` in the entropy data frame. The tract diversity index is visualized using **ggplot2** relative to its travel time to downtown Los Angeles, with a LOESS smoother superimposed over the scatterplot to represent the diversity gradient.

```{r}
la_entropy_geo$minutes <- as.numeric(mins_to_dwntwn)

ggplot(la_entropy_geo, aes(x = minutes, y = entropy)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "loess") + 
  theme_minimal() + 
  scale_x_continuous(limits = c(0, 80)) + 
  labs(title = "Diversity gradient, Los Angeles urbanized area",
       x = "Travel-time to downtown Los Angeles in minutes, Census tracts",
       y = "Entropy index")
```

The visualization of the diversity gradient shows that neighborhood diversity increases with driving time from the urban core in Los Angeles, peaking at about 35 minutes in free-flowing traffic from the urban core then leveling off after that. The structure of the diversity gradient suggests that Census tracts near to downtown tend to be segregated, and suburban tracts more likely to be integrated.

